---
title: "Integrating spectral flow and CyTOF data"
author: "Christina Bligaard Pedersen"
date: "September 14, 2021"
output: 
  html_document:
    toc: true
    toc_float: true
    highlight: kate
    df_print: paged
---
<style type="text/css">

h1.title {
  color: #004d66;
}
h4.author {
  font-style: italic;
  font-size: 18px;
  color: #008cba;
}
h4.date {
  font-style: italic;
  font-size: 16px;
  color: #008cba;
}
h1 { /* Header 1 */
  color: #004d66;
  font-size: 28px;
}
h2 { /* Header 2 */
  color: #004d66;
  font-size: 22px;
}
h3 { /* Header 3 */
  color: #004d66;
  font-size: 16px;
}
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>


<br>

This vignette will demonstrate the integration of spectral flow cytometry (SFC) and CyTOF protein expression measurements using cyCombine.

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  strip.white = T, comment = ""
)

knitr::opts_knit$set(root.dir = '/home/projects/dp_immunoth/people/s134891/cycombine/flow_cytof/')

```

<br>

<!-- For this purpose, we will utilize a paired dataset from [Ferrer-Font et al. (2020)](https://doi.org/10.1002/cyto.a.24016). This group studied the comparability of CyTOF and SFC data by using splenocytes from the same samples which had been quantified using the two technologies. They were able to conclude that the results were indeed very similar. However, this analysis was carried out by analyzing the sets separately. Here, we would like to demonstrate the *direct* integration of the sets. -->

<!-- According to the article, the SFC panel of this study contained 21 protein markers, and the CyTOF panel comprised 36 markers. 20 of these were overlapping. A total of five samples are included for each technology, and they are available from FlowRepository (ID: [FR-FCM-Z2GZ](https://flowrepository.org/id/FR-FCM-Z2GZ)) or [GitHub](https://github.com/SameOldSamOld/CyTOFAuroraComparison). All files were downsampled to 19,858 cells in the analysis of the article and the uploaded FCS files. We downloaded these and relied on one of the GitHub scripts for accurate mapping of channels and markers. -->


<!-- In the original study, they focused on 11 immune cell populations, which were manually gated. Some of these assignments overlap - i.e. T cells were considered as a combined population and after gating on CD4/CD8. The included populations were "T cells", "CD4 T cells", "CD8 T cells", "DN T cells", "B cells", "NK cells", "Monocytes", "DCs", "DC1s", "DC2s", and "Neutrophils". However, the authors have not shared the cell labels in their FCS files, so we did not dig into these details here. -->

<!-- It was already established that the frequencies of these populations correlated nicely between the two technologies, however the article also states that they were unable to overlay the two data types due to the different data structures of the platforms. With this in mind, we enter the path of cyCombine. -->


<br>

In this vignette, we will analyze the healthy donor PBMC SFC and CyTOF data, which is also presented in the [three-platform vignette](https://biosurf.org/cyCombine_CITEseq_Spectral_CyTOF.html).

The SFC data is from [Park et al. (2020)](https://doi.org/10.1002/cyto.a.24213) available from FlowRepository (ID: [FR-FCM-Z2QV](https://flowrepository.org/id/FR-FCM-Z2QV)). We pre-gated to live single cells in FlowJo version 10 (Tree Star Inc). Singlets and non-debris were identified using forward and side-scatter. Dead cells were excluded using live/dead stains. Data from these gates were then exported in FCS format.

For the CyTOF data, we use the data from a single healthy donor processed at the Human Immune Monitoring Center. The sample was also derived from FlowRepository (ID: [FR-FCM-ZYAJ](https://flowrepository.org/id/FR-FCM-ZYAJ)) and pre-gated to live intact singlets in FlowJo version 10 (Tree Star Inc). 

<br>


We start by loading some packages:

```{r libraries, results = 'hide', warning=FALSE, message=FALSE}
library(cyCombine)
library(tidyverse)

# Set a seed for reproducibility
seed <- 9369

```


<br>

<!-- # Ferrer-Font dataset (same samples) -->

<!-- ## Loading data -->
<!-- ### Spectral flow cytometry data -->
<!-- We are now ready to load the spectral flow data. We have set up a panel file in csv format, so the correct information is extractable from there. Let us have a look at the contents: -->

<!-- ```{r loading spectral flow data 1, message=F, warning=F} -->
<!-- # Directory with raw .fcs files -->
<!-- data_dir <- "FR-FCM-Z2GZ/SFC/" -->

<!-- # Panel and reading data -->
<!-- sfc_panel <- read_csv(paste0(data_dir, "/panel_sfc.csv")) -->
<!-- sfc_panel -->

<!-- ``` -->

<!-- <br> -->
<!-- We then progress with reading the spectral flow dataset and converting it to a tibble format, which is easy to process. We use cofactor = 6000 for spectral flow data asinh-transformation as suggested by the original study. Because the FCS files do not contain the marker descriptions we specify these based on the `sfc_panel` object we prepared. -->

<!-- ```{r loading spectral flow data 2} -->
<!-- # Extract the markers to keep -->
<!-- sfc_markers <- sfc_panel %>%  -->
<!--   dplyr::filter(Type != "none") %>%  -->
<!--   pull(Marker) %>%  -->
<!--   str_remove_all("[ _-]") -->

<!-- # Read fcs files and convert to tibble -->
<!-- spectral <- prepare_data(data_dir = data_dir, -->
<!--                          metadata = paste0(data_dir, "../metadata.xlsx"), -->
<!--                          sample_ids = 'SampleID', -->
<!--                          filename_col = "FileName", -->
<!--                          batch_ids = "Batch", -->
<!--                          cofactor = 6000, -->
<!--                          derand = FALSE, -->
<!--                          markers = sfc_markers, -->
<!--                          down_sample = FALSE, -->
<!--                          panel = sfc_panel, -->
<!--                          panel_channel = 'Channel', -->
<!--                          panel_antigen = 'Marker') -->

<!-- ``` -->


<!-- <br><br> -->



<!-- ### CyTOF data -->
<!-- We are now ready to load the CyTOF data. Once again, let us have a look at the panel - we also need it here to make our file reading a bit easier. -->

<!-- ```{r loading cytof data 1, message=F, warning=F} -->
<!-- # Directory with raw .fcs files -->
<!-- data_dir <- "FR-FCM-Z2GZ/CyTOF/" -->

<!-- # Panel and reading data -->
<!-- cytof_panel <- read_csv(paste0(data_dir, "/panel_cytof.csv")) -->
<!-- cytof_panel -->

<!-- ``` -->

<!-- <br> -->

<!-- And then for the FCS reading and conversion to tibble format. We will employ derandomization and asinh-trasnformation with cofactor = 5. -->

<!-- ```{r loading cytof data 2} -->
<!-- # Extract the markers to keep -->
<!-- cytof_markers <- cytof_panel %>%  -->
<!--   dplyr::filter(Type != "none") %>%  -->
<!--   pull(Marker) %>%  -->
<!--   str_remove_all("[ _-]") -->

<!-- # Read fcs files and convert to tibble -->
<!-- cytof <- prepare_data(data_dir = data_dir, -->
<!--                       metadata = paste0(data_dir, "../metadata.xlsx"), -->
<!--                       sample_ids = 'SampleID', -->
<!--                       filename_col = "FileName", -->
<!--                       batch_ids = "Batch", -->
<!--                       cofactor = 5, -->
<!--                       derand = TRUE, -->
<!--                       markers = cytof_markers, -->
<!--                       down_sample = FALSE, -->
<!--                       panel = cytof_panel, -->
<!--                       panel_channel = 'Channel', -->
<!--                       panel_antigen = 'Marker') -->

<!-- ``` -->



<!-- <br><br> -->


<!-- ## Batch correction -->

<!-- Next, we need to figure out which markers to use when integrating the two datasets. From the publication, we know that the overlap is 20 markers - let's verify that to make sure naming matches 1:1. -->

<!-- ```{r overlapping markers} -->
<!-- intersect(sfc_markers, cytof_markers) -->

<!-- ``` -->

<!-- <br> -->

<!-- Now, we are ready to combine the two datasets on the overlapping columns. -->

<!-- ```{r combine sets} -->
<!-- # Get overlapping column names -->
<!-- overlap_cols <- intersect(colnames(spectral), colnames(cytof)) -->

<!-- # Make one tibble -->
<!-- uncorrected <- bind_rows(cytof[,overlap_cols], spectral[,overlap_cols]) %>% -->
<!--   mutate(id = 1:(nrow(cytof)+nrow(spectral))) -->

<!-- ``` -->

<!-- And now, batch correction can be performed with cyCombine. We use rank for normalization because these are data from two different technologies so they are likely to have more different distributions, and ranking is more appropriate in such cases. -->

<!-- ```{r batch correction, message=FALSE, warning=FALSE, error=FALSE, results='hide'} -->
<!-- # Run batch correction  -->
<!-- corrected <- uncorrected %>% -->
<!--   batch_correct(xdim = 8, -->
<!--                 ydim = 8, -->
<!--                 norm_method = 'rank', -->
<!--                 ties.method = 'average', -->
<!--                 seed = seed) -->

<!-- ``` -->

<!-- <br> -->


<!-- ## Evaluating performance -->

<!-- We now evaluate the correction using EMD - each marker is evaluated across all cells. -->

<!-- ```{r global emd, message=FALSE, fig.height=6, fig.width=12} -->
<!-- corrected$label <- uncorrected$label <- 1 # We set all labels to 1 in order to evaluate globally -->

<!-- emd <- evaluate_emd(uncorrected = uncorrected, -->
<!--                     corrected = corrected) -->

<!-- cowplot::plot_grid(emd$violin, emd$scatterplot) -->
<!-- ``` -->
<!-- <br> -->

<!-- We also use the MAD score for evaluation. -->
<!-- ```{r mad, message=FALSE} -->

<!-- # MAD -->
<!-- mad <- uncorrected %>% -->
<!--   cyCombine::evaluate_mad(corrected, -->
<!--                           filter_limit = NULL) -->

<!-- cat('The MAD score is: ', mad$score, '\n') -->

<!-- ``` -->


<!-- <br> -->


<!-- Finally, let us look at some plots to visualize the correction. First, the marker distributions before and after: -->

<!-- ```{r density plot, message=FALSE, fig.height=20, fig.width=16} -->
<!-- plot_density(uncorrected, corrected, ncol = 4) -->

<!-- ``` -->

<!-- <br> -->
<!-- Finally, some UMAPs to visualize the correction. I will downsample to 10,000 cells from each dataset so it is easier to see what is going on. -->

<!-- ```{r umaps, fig.height=8, fig.width=16} -->

<!-- inds <- split(1:length(uncorrected$batch), uncorrected$batch) -->
<!-- sample <- unlist(lapply(inds, sample, 10000)) -->

<!-- plot1 <- plot_dimred(uncorrected[sample,], name = 'Uncorrected', type = 'umap') -->
<!-- plot2 <- plot_dimred(corrected[sample,], name = 'Corrected', type = 'umap') -->

<!-- cowplot::plot_grid(plot1, plot2) -->

<!-- ``` -->

<!-- <br> -->

<!-- When studying the plots it is apparent that the datasets from the two platforms are much more comparable after batch correction. The two data types occupy the same regions of the UMAP to a much larger extent post-correction.  -->

<br><br>


# Loading data

We start by defining some colors to use.

```{r colors, results = 'hide', warning=FALSE, message=FALSE}
# Define some nice colors for plotting
color_clusters <- c("#DC050C", "#FB8072", "#1965B0", "#7BAFDE", "#882E72",
                    "#B17BA6", "#FF7F00", "#FDB462", "#E7298A", "#E78AC3",
                    "#33A02C", "#B2DF8A", "#55A1B1", "#8DD3C7", "#A6761D",
                    "#E6AB02", "#7570B3", "#BEAED4", "#666666", "#999999",
                    "#aa8282", "#d4b7b7", "#8600bf", "#ba5ce3", "#808000",
                    "#aeae5c", "#1e90ff", "#00bfff", "#FAE174", "#56ff0d",
                    "#ffff00", "#D4E1C8", "#D470C8", "#64C870", "#0AC8D4",
                    "#64C80C", "#641400", "#000000", "#0C00FA", "#7800FA",
                    "#FA00FA", "#00FAFA", "#707A00", "#0C0C91", "#B5651D")
```

<br>

## Spectral flow cytometry data pre-processing

We are now ready to load the spectral flow data into a tibble.

```{r loading spectral flow data, paged.print = T, message=F, warning=F, results='hide'}
# Directory with raw .fcs files
data_dir <- "Park_et_al_2020_Live+/"

# Panel and reading data
sfc_panel <- read_csv(paste0(data_dir, "/panel_Park2020.csv"))

sfc_markers <- sfc_panel %>%
  dplyr::filter(Type != "none") %>%
  pull(Marker) %>%
  str_remove_all("[ _-]")

spectral <- prepare_data(data_dir = data_dir,
                         metadata = paste0(data_dir, "/Spectral samples cohort.xlsx"),
                         filename_col = "FCS_name",
                         batch_ids = "Batch",
                         condition = "Set",
                         sample_ids = "Patient id",
                         cofactor = 6000,
                         derand = FALSE,
                         markers = sfc_markers,
                         down_sample = FALSE)


# Subset to a single sample for simplicity
spectral <- spectral %>%
  dplyr::filter(sample == "303444")

```

Now, we a single sample consisting of 582,005 cells. We now want to generate some cell labels using the overlapping markers

```{r label sfc, message=F, warning=F, results = 'hide', fig.show='hide'}
# Defining the set of overlapping markers - there are 26
overlap_markers <- c('CD3', 'CD4', 'CD8', 'CD25', 'CCR7', 'CD45RA', 'TCRgd', 'CD27', 'CD19', 'CD20', 'CD24', 'IgD', 'CXCR5', 'CD56', 'CD57', 'CD16', 'CD38', 'CD28', 'CCR6', 'CD127', 'HLADR', 'CD14', 'CD11c', 'CD123', 'CXCR3', 'PD1')

# Time to label the cells. Here, we will apply a SOM and ConsensusClusterPlus

# Clustering with kohonen 10x10
set.seed(seed)
som_ <- spectral %>%
  dplyr::select(dplyr::all_of(overlap_markers)) %>%
  as.matrix() %>%
  kohonen::som(grid = kohonen::somgrid(xdim = 10, ydim = 10),
               rlen = 10,
               dist.fcts = "euclidean")


cell_clustering_som <- som_$unit.classif
codes <- som_$codes[[1]]

# Meta-clustering with ConsensusClusterPlus
mc <- ConsensusClusterPlus::ConsensusClusterPlus(t(codes), maxK = 35, reps = 100,
                                                 pItem = 0.9, pFeature = 1, plot = F,
                                                 clusterAlg = "hc", innerLinkage = "average", 
                                                 finalLinkage = "average",
                                                 distance = "euclidean", seed = seed)


# Get cluster ids for each cell - here we choose to look at 30 meta-clusters but we will merge some in the next step
code_clustering1 <- mc[[30]]$consensusClass
spectral$label <- as.factor(code_clustering1[cell_clustering_som])

```


```{r UMAP spectral}
# Down-sampling for plot (20,000 cells)
set.seed(seed)
spectral_sliced <- spectral %>%
  dplyr::slice_sample(n = 20000)

# Let us visualize these clusters on a UMAP
umap <- spectral_sliced %>% plot_dimred(name = 'SFC clusters')
spectral_sliced <- cbind.data.frame(umap$data, spectral_sliced)

ggplot(spectral_sliced, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = label), alpha = 0.3, size = 0.4) +
  theme_bw() + theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = color_clusters) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 1)))

```

```{r violin plots spectral}
# And now let us use violin plots for the assignment of labels for each population
p <- list()
for (m in overlap_markers) {
  p[[m]] <- ggplot(spectral, aes_string(x = 'label', y = m)) +
    geom_violin(aes(fill = label)) +
    scale_fill_manual(values = color_clusters) +
    theme_bw()
}

p2 <- ggpubr::ggarrange(plotlist = p, common.legend = T)

```

```{r violin plots spectral 2, fig.height=22, fig.width=35}
p2
```

Based on these plots and the UMAP, it is possible to define labels for many of the clusters, although some also appear strange. This includes the very small cluster 2, which could be B-T cell doublets (CD19+CD3+). Cluster 22 was also very small and expressed only CD25, CD127, CD45RA. Cluster 24 also seemed very mixed with bimodal CD14 and CD11c distributions. Finally cluster 29 had only 3 cells, and was left unlabeled.



<br>

Regarding cluster 10, this was also tricky, but considering both its UMAP location and the intermediate level of CD4, which is comparable to clusters 1, 2, and 5, we are comfortable with labeling these as myeloid cells. The rest of the labels are assigned below.

```{r label sfc naming}
spectral_cl_names <- c('1' = 'B cells', '2' = 'Unlabeled', '3' = 'Naive CD8+ T cells', '4' = 'Memory CD8+ T cells', '5' = 'gdT cells',
                       '6' = 'Memory CD4+ T cells', '7' = 'Naive CD4+ T cells', '8' = 'B cells (IgDlo)', '9' = 'gdT cells (CD57dim)', '10' = 'Memory CD8+ T cells (CD57+)',
                       '11' = 'NK cells (CD16+CD57+)', '12' = 'NK cells (CD16+CD57-)', '13' = 'NK cells (CD16-CD57-)', '14' = 'NK cells (CD16-CD57+)', '15' = 'Memory CD4+ T cells (CD57+ PD1+)',
                       '16' = 'NK cells (CD16-CD57-)', '17' = 'pDCs', '18' = 'Tregs', '19' = 'pDCs', '20' = 'B cells (CD19lo CD38++)',
                       '21' = 'CD3lo gdT cells' , '22' = 'Unlabeled', '23' = 'pDCs (CD38+)', '24' = 'Unlabeled', '25' = 'Memory CD4+ T cells',
                       '26' = 'mDCs' , '27' = 'Classical monocytes', '28' = 'Non-classical monocytes', '29' = 'Unlabeled', '30' = 'Classical monocytes') 

# c('1' = '', '2' = '', '3' = '', '4' = '', '5' = '',
#                        '6' = '', '7' = '', '8' = '', '9' = '', '10' = '',
#                        '11' = '', '12' = '', '13' = '', '14' = '', '15' = '',
#                        '16' = '', '17' = '', '18' = '', '19' = '', '20' = '',
#                        '21' = '' , '22' = '', '23' = '', '24' = '', '25' = '',
#                        '26' = '' , '27' = '', '28' = '', '29' = '', '30' = '')


spectral$label <- spectral_cl_names[spectral$label]

# Removing the unlabeled cells
spectral <- spectral %>%
  dplyr::filter(label != 'Unlabeled')

```

After removal of the unlabeled cells, we have 573,397 cells remaining (98.5 %) in the SFC dataset and this portion of the data is now ready for batch correction. We will now look at the CyTOF data.




<br>

# CyTOF data pre-processing

Then it is time to read the CyTOF data. We use a single sample (ctrls-001) from FlowRepository: FR-FCM-ZYAJ. We downloaded the version normalized with MATLAB and pre-gated it to live intact singlets using FlowJo.

```{r cytof read, message=F, warning=F, results='hide'}
### Get HIMC cytof data for one sample
# Panel and reading data
cytof_panel <- read_csv("cytof_panel_HIMC.csv")

cytof_markers <- cytof_panel %>%
  dplyr::filter(Type != "none") %>%
  pull(Marker) %>%
  str_remove_all("[ _-]")

# Read the fcs file and make it a dataframe
cytof <- prepare_data(data_dir = '.',
                      cofactor = 5,
                      derand = T,
                      markers = cytof_markers,
                      down_sample = FALSE,
                      batch_ids  = c('CyTOF'),
                      sample_ids = c('Sample1'))

```



Now, we a single sample consisting of 174,601 cells. Similarly to the other datasets, we now need to generate some cell labels - based on the overlapping markers only. Let us look at this:

```{r label cytof, message=F, warning=F, results = 'hide', fig.show='hide'}
# Subset to overlap
cytof <- cytof %>%
  select(all_of(c(overlap_markers, 'sample', 'batch')))

# Time to label the cells. Here, we will again apply a SOM and ConsensusClusterPlus

# Clustering with kohonen
set.seed(seed)
som_ <- cytof %>%
  dplyr::select(dplyr::all_of(overlap_markers)) %>%
  as.matrix() %>%
  kohonen::som(grid = kohonen::somgrid(xdim = 10, ydim = 10),
               rlen = 10,
               dist.fcts = "euclidean")


cell_clustering_som <- som_$unit.classif
codes <- som_$codes[[1]]

# Meta-clustering with ConsensusClusterPlus
mc <- ConsensusClusterPlus::ConsensusClusterPlus(t(codes), maxK = 35, reps = 100,
                                                 pItem = 0.9, pFeature = 1, plot = F,
                                                 clusterAlg = "hc", innerLinkage = "average", 
                                                 finalLinkage = "average",
                                                 distance = "euclidean", seed = seed)


# Get cluster ids for each cell - here we also look at 30 meta-clusters but we will merge some in the next step
code_clustering1 <- mc[[30]]$consensusClass
cytof$label <- as.factor(code_clustering1[cell_clustering_som])

```

```{r UMAP cytof}
# Down-sampling for plot (20,000 cells)
set.seed(seed)
cytof_sliced <- cytof %>%
  dplyr::slice_sample(n = 20000)

# Let us visualize these clusters on a UMAP
umap <- cytof_sliced %>% plot_dimred(name = 'CyTOF clusters')
cytof_sliced <- cbind.data.frame(umap$data, cytof_sliced)

ggplot(cytof_sliced, aes(x = UMAP1, y = UMAP2)) +
  geom_point(aes(color = label), alpha = 0.3, size = 0.4) +
  theme_bw() + theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = color_clusters) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 1)))

```


```{r violin plots cytof}
# And now let us use violin plots for the assignment of labels for each population
p <- list()
for (m in overlap_markers) {
  p[[m]] <- ggplot(cytof, aes_string(x = 'label', y = m)) +
    geom_violin(aes(fill = label)) +
    scale_fill_manual(values = color_clusters) +
    theme_bw()
}

p2 <- ggpubr::ggarrange(plotlist = p, common.legend = T)

```

```{r violin plots cytof 2, fig.height=22, fig.width=35}
p2
```


<br>
Now we assign labels to each of the clusters. There are some labels which we did not have for the SFC data.

```{r label cytof naming}
cytof_cl_names <- c('1' = 'B cells', '2' = 'Naive CD4+ T cells', '3' = 'Memory CD4+ T cells (CD38+)', '4' = 'Memory CD4+ T cells', '5' = 'B cells (CD19lo CD38++)',
                       '6' = 'B cells (IgDlo)', '7' = 'Dbl. neg. T cells', '8' = 'Dbl. neg. T cells', '9' = 'Memory CD8+ T cells (CD57+)', '10' = 'Naive CD8+ T cells',
                       '11' = 'Memory CD8+ T cells', '12' = 'Tregs', '13' = 'CD8+ TEMRA cells', '14' = 'gdT cells', '15' = 'Tregs',
                       '16' = 'NK cells (CD16-CD57+)', '17' = 'gdT cells (CD57dim)', '18' = 'NK cells (CD16-CD57-)', '19' = 'Memory CD8+ T cells', '20' = 'Memory CD8+ T cells (CD38+)',
                       '21' = 'NK cells (CD16+CD57+)' , '22' = 'CD8+ TEMRA cells', '23' = 'Memory CD4+ T cells (CD57+)', '24' = 'mDCs', '25' = 'Classical monocytes',
                       '26' = 'NK cells (CD16+CD57-)' , '27' = 'pDCs', '28' = 'Non-classical monocytes', '29' = 'mDCs', '30' = 'mDCs')


cytof$label <- cytof_cl_names[cytof$label]

```





<br><br>


# Batch correction

Now, we are ready to combine the two datasets on the overlapping columns. As discussed above, there are 26 markers which overlap between the sets. In this example, we do not downsample to have the same number of cells from each platform, but instead analyze all available cells.

```{r combine sets 2}
# Get overlapping column names
overlap_cols <- intersect(colnames(cytof), colnames(spectral))

# Make one tibble - here, we downsample to have the same number of cells from each platform
uncorrected <- bind_rows(cytof[,overlap_cols], spectral[,overlap_cols]) %>%
  mutate(id = 1:(nrow(cytof)+nrow(spectral)))

```

<br>

Now, batch correction can be performed with cyCombine. Because these are datasets of completely different platforms, we use rank as the normalization method - similar to what is done for the three-platform integration including CITE-seq data. In addition, we have also selected to use a 3x3 grid for the batch correction, as this looked much better when inspecting the density plots.

```{r batch correction, message=FALSE, warning=FALSE, error=FALSE, results='hide'}
# Run batch correction
corrected <- uncorrected %>%
  batch_correct(seed = seed,
                xdim = 3,
                ydim = 3,
                norm_method = 'rank',
                ties.method = 'average')

```


```{r labels to factor}
# Let's make a meaningful ordering of the labels
label_levels <- c('B cells', 'B cells (CD19lo CD38++)', 'B cells (IgDlo)', 'Naive CD4+ T cells', 'Memory CD4+ T cells', 'Memory CD4+ T cells (CD57+)', 'Memory CD4+ T cells (CD57+ PD1+)', 'Memory CD4+ T cells (CD38+)', 'Tregs', 'Naive CD8+ T cells',  'Memory CD8+ T cells', 'Memory CD8+ T cells (CD57+)', 'Memory CD8+ T cells (CD38+)', 'CD8+ TEMRA cells', 'Dbl. neg. T cells', 'gdT cells', 'gdT cells (CD57dim)', 'CD3lo gdT cells', 'NK cells (CD16-CD57-)', 'NK cells (CD16-CD57+)', 'NK cells (CD16+CD57-)', 'NK cells (CD16+CD57+)', 'Classical monocytes', 'Non-classical monocytes', 'mDCs', 'pDCs', 'pDCs (CD38+)')

# Add labels back from uncorrected
uncorrected$label <- factor(uncorrected$label, levels = label_levels)
corrected$label <- uncorrected$label
```



<br><br>

# Evaluating batch correction

We can now evaluate the correction using the EMD reduction - first, we apply clustering and then evaluate each marker in each cluster.

```{r emd, message=FALSE, fig.height=6, fig.width=12}
# Generate labels for EMD evaluation
labels <- corrected %>%
  create_som(seed = seed,
             xdim = 8,
             ydim = 8,
             markers = get_markers(corrected))

# Add labels
corrected <- corrected %>%
  dplyr::mutate(som = labels)
celltype_col <- "som"

uncorrected <- corrected %>%
  dplyr::select(id, all_of(celltype_col)) %>%
  dplyr::left_join(uncorrected, by = "id")


# Evaluate Earth Movers Distance
emd <- uncorrected %>%
  evaluate_emd(corrected, markers = get_markers(corrected), cell_col = celltype_col)

# Show plots
cowplot::plot_grid(emd$violin, emd$scatterplot)

```


<br>

We also use the MAD score for evaluation:

```{r mad, message=FALSE}
# MAD
mad <- uncorrected %>%
  cyCombine::evaluate_mad(corrected,
                          markers = get_markers(corrected),
                          cell_col = celltype_col,
                          filter_limit = NULL)

cat('The MAD score is: ', mad$score, '\n')

```


<br>

For this integration, the EMD reduction is 0.72 and the MAD score is 0.05, which are very satisfactory values.

<br>

However, as one should always do we will also visualize the correction with plots. First, the marker distributions before and after:

```{r density plot, message=FALSE, fig.height=12, fig.width=16}
plot_density(uncorrected, corrected, ncol = 4, xlims = c(-2, 9))

```

Let us also see the UMAPs for uncorrected and corrected data colored by batch. We downsample a bit here so it's easier to see what is going on.

```{r umaps 1, fig.height=8, fig.width=16}
# Down-sampling for plot (20,000 cells from each platform)
set.seed(seed)
uncorrected_sliced <- uncorrected %>%
  dplyr::group_by(batch) %>%
  dplyr::slice_sample(n = 20000) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(id)

corrected_sliced <- corrected %>%
  dplyr::filter(id %in% uncorrected_sliced$id)


# UMAP plot uncorrected
umap1 <- uncorrected_sliced %>%
  plot_dimred(name = "uncorrected", type = "umap")

# UMAP plot corrected
umap2 <- corrected_sliced %>%
  plot_dimred(name = "corrected", type = "umap")

# Show plots
cowplot::plot_grid(umap1, umap2)

```

Now, we also have labels for each of the datasets, which were generated independently of the batch correction. Let us have some visualization with these.


```{r umaps labels, fig.height=9, fig.width=12}
# For uncorrected data, let's make a combined object
uncor_df <- cbind.data.frame(umap1$data[,1:2], uncorrected_sliced)

umap3 <- ggplot(uncor_df, aes(x = UMAP1, y = UMAP2)) +
                geom_point(aes(color = label), alpha = 0.3, size = 0.4) +
                theme_bw() + theme(plot.title = element_text(hjust = 0.5)) +
                ggtitle('UMAP - uncorrected') +
                scale_color_manual(values = color_clusters) + facet_wrap(~batch) +
                guides(color = guide_legend(override.aes = list(alpha = 1, size = 1)))

# For corrected data, let's make a combined object
cor_df <- cbind.data.frame(umap2$data[,1:2], corrected_sliced)

umap4 <- ggplot(cor_df, aes(x = UMAP1, y = UMAP2)) +
                geom_point(aes(color = label), alpha = 0.3, size = 0.4) +
                theme_bw() + theme(plot.title = element_text(hjust = 0.5)) +
                ggtitle('UMAP - corrected') +
                scale_color_manual(values = color_clusters) + facet_wrap(~batch) +
                guides(color = guide_legend(override.aes = list(alpha = 1, size = 1)))


# Show plots
cowplot::plot_grid(umap3, umap4, nrow = 2)

```

<br>

We can also show the UMAPs faceted by technology, where cells are colored by their expression of the 26 overlapping markers:


```{r umaps markers split, fig.height=70, fig.width=10}
# Loop over markers to make the plots for both the uncorrected and corrected data
expr_plots <- list()
for (m in get_markers(corrected)) {

  # Find range for marker to make plots comparable
  range <- summary(c(uncor_df[,m], cor_df[,m]))[c(1,6)]

  expr_plots[[paste0(m, '1')]] <- ggplot(uncor_df, aes(x = UMAP1, y = UMAP2)) +
                geom_point(aes_string(color = m), alpha = 0.3, size = 0.4) +
                theme_bw() + theme(plot.title = element_text(hjust = 0.5)) +
                ggtitle('UMAP - uncorrected') +
                scale_color_gradientn(m, limits  = range, colours = colorRampPalette(rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral")))(50)) +
                facet_wrap(~batch)

  expr_plots[[paste0(m, '2')]] <- ggplot(cor_df, aes(x = UMAP1, y = UMAP2)) +
                geom_point(aes_string(color = m), alpha = 0.3, size = 0.4) +
                theme_bw() + theme(plot.title = element_text(hjust = 0.5)) +
                ggtitle('UMAP - corrected') +
                scale_color_gradientn(m, limits = range, colours = colorRampPalette(rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral")))(50)) +
                facet_wrap(~batch)

}

# Show plots
cowplot::plot_grid(plotlist = expr_plots, ncol = 2)

```


<br>

To determine if it is possible to obtain a similar cell labeling *after* correction, we will now re-cluster and label the combined, corrected dataset.


```{r label combined, message=F, warning=F, results = 'hide', fig.show='hide'}
# Clustering with kohonen
set.seed(seed)
som_ <- corrected %>%
  dplyr::select(dplyr::all_of(overlap_markers)) %>%
  as.matrix() %>%
  kohonen::som(grid = kohonen::somgrid(xdim = 10, ydim = 10),
               rlen = 10,
               dist.fcts = "euclidean")


cell_clustering_som <- som_$unit.classif
codes <- som_$codes[[1]]

# Meta-clustering with ConsensusClusterPlus
mc <- ConsensusClusterPlus::ConsensusClusterPlus(t(codes), maxK = 45, reps = 100,
                                                 pItem = 0.9, pFeature = 1, plot = F,
                                                 clusterAlg = "hc", innerLinkage = "average", 
                                                 finalLinkage = "average",
                                                 distance = "euclidean", seed = seed)


# Get cluster ids for each cell - here we will look at 40 meta-clusters to attempt to grasp the full range of cell types from both data sources but we will merge some in the next step
code_clustering1 <- mc[[35]]$consensusClass
corrected$new_label <- as.factor(code_clustering1[cell_clustering_som])

```


```{r violin plots combined}
# And now let us use violin plots for the assignment of labels for each population
p <- list()
for (m in overlap_markers) {
  p[[m]] <- ggplot(corrected, aes_string(x = 'new_label', y = m)) +
    geom_violin(aes(fill = new_label)) +
    scale_fill_manual(values = color_clusters) +
    theme_bw()
}

p2 <- ggpubr::ggarrange(plotlist = p, common.legend = T)

```

```{r violin plots combined 2, fig.height=22, fig.width=35}
p2
```


<br>
Now we assign labels to each of the clusters. There are some labels which we did not have for the separate data sets, 

```{r label combined naming}
corrected_cl_names <- c('1' = 'NK cells (CD16+CD57+)', '2' = 'NK cells (CD16+CD57-)', '3' = 'Naive CD8+ T cells', '4' = 'CD3lo gdT cells', '5' = 'Naive CD4+ T cells',
                        '6' = 'Memory CD8+ T cells (CD57+)', '7' = 'Memory CD8+ T cells', '8' = 'B cells (IgDlo)', '9' = 'B cells', '10' = 'NK cells (CD16-CD57-)',
                        '11' = 'gdT cells (CD57dim)', '12' = 'Dbl. neg. T cells', '13' = 'pDCs', '14' = 'NK cells (CD16-CD57+)', '15' = 'Memory CD8+ T cells (CD57+)',
                        '16' = 'Memory CD4+ T cells (CD57+)', '17' = 'B cells', '18' = 'pDCs (CD38+)', '19' = 'gdT cells', '20' = 'B cells (CD19lo CD38++)',
                        '21' = 'mDCs' , '22' = 'Memory CD8+ T cells (CD38+)', '23' = 'Tregs', '24' = 'Memory CD4+ T cells', '25' = 'Memory CD4+ T cells (CD38+)',
                        '26' = 'mDCs' , '27' = 'Memory CD8+ T cells (CD38+)', '28' = 'Memory CD4+ T cells', '29' = 'Memory CD4+ T cells', '30' = 'Non-classical monocytes',
                        '31' = 'Non-classical monocytes', '32' = 'Classical monocytes', '33' = 'Classical monocytes', '34' = 'Doublets?', '35' = 'Non-classical monocytes')

# Add the new label to the levels
label_levels <- c(label_levels, 'Doublets?')

# Add labels
corrected$new_label <- factor(corrected_cl_names[corrected$new_label], levels = label_levels)
```

<br> 
Now that we have these labels, we can plot the corrected UMAP with them.


```{r umap new labels, fig.height=4.5, fig.width=12}
# For corrected data, let's make a combined object
corrected_sliced <- corrected %>%
  dplyr::filter(id %in% uncorrected_sliced$id)

cor_df <- cbind.data.frame(umap2$data[,1:2], corrected_sliced)

umap5 <- ggplot(cor_df, aes(x = UMAP1, y = UMAP2)) +
                geom_point(aes(color = new_label), alpha = 0.3, size = 0.4) +
                theme_bw() + theme(plot.title = element_text(hjust = 0.5)) +
                ggtitle('UMAP - corrected - reclustered') +
                scale_color_manual(values = color_clusters[c(1:6,8:12,14:30)]) + facet_wrap(~batch) +
                guides(color = guide_legend(override.aes = list(alpha = 1, size = 1)))

# Show plot
umap5

```


<br> 

Finally, let us compare the fractions of the 'corrected' labels with those from the uncorrected, separately labeled sets.

```{r compare label props}
# Split corrected data in to original parts
corrected_spectral <- corrected %>% 
  dplyr::filter(batch == 'Spectral')

corrected_cytof <- corrected %>% 
  dplyr::filter(batch == 'CyTOF')


# Compare fractions
Spectral_uncor <- (table(factor(spectral$label, levels = label_levels)) / nrow(spectral) ) * 100
Spectral_cor <- (table(factor(corrected_spectral$new_label, levels = label_levels)) / nrow(spectral) ) * 100

CyTOF_uncor <- (table(factor(cytof$label, levels = label_levels)) / nrow(cytof) ) * 100
CyTOF_cor <- (table(factor(corrected_cytof$new_label, levels = label_levels)) / nrow(cytof) ) * 100



# Barplots
ggdf <- reshape2::melt(cbind(Spectral_uncor, Spectral_cor, CyTOF_uncor, CyTOF_cor))
colnames(ggdf) <- c('Cluster', 'Set', 'Percentage')
ggdf$Cluster <- factor(ggdf$Cluster, levels = label_levels)

ggplot(ggdf, aes(x = Set, y = Percentage, fill = Cluster)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  scale_fill_manual(values = color_clusters) +
  ggtitle('Clustering results') +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab('Percentage of total cells') + xlab('') + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


```


<br>

From this comparison, it is clear that most proportions of the uncorrected-and-labeled sets are well-maintained in the corrected-and-labeled set. There are some discrepancies, but some of those are also found for labels, where there was not necessarily a 100 % clear distinction between + and - (e.g. for NK cells and their expression of CD57). We also note that even while some populations are not found for the corrected set, that does not mean that those cells would not be identifiable if using a higher number of meta-clusters.

In conclusion, even more detailed clusters are well-preserved after batch correction with cyCombine.

