---
title: "Integrating CITE-seq - ADT and RNA"
author: "Søren H Dam"
date: "Last updated on `r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    highlight: kate
    df_print: paged
---

<style type="text/css">

h1.title {
  color: #004d66;
}
h4.author {
  font-style: italic;
  font-size: 18px;
  color: #008cba;
}
h4.date {
  font-style: italic;
  font-size: 16px;
  color: #008cba;
}
h1 { /* Header 1 */
  color: #004d66;
  font-size: 28px;
}
h2 { /* Header 2 */
  color: #004d66;
  font-size: 22px;
}
h3 { /* Header 3 */
  color: #004d66;
  font-size: 16px;
}
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# rmarkdown::render("cyCombine_CITEseq_ADT_RNA.Rmd")
```

This vignette demonstrates integrating the RNA and ADT parts of a CITE-seq dataset using cyCombine.
In order to assimilate the two data modalities, we use our own implementation of a Single-cell nearest-neighbor pseudobulking ([scennep](https://github.com/shdam/scennep)).

## Dependencies

```{r dependencies}
# remotes::install_github("mojaveazure/seurat-disk#198")
library("Seurat")
library("SeuratDisk")
library("dplyr")
library("ggplot2")
library("patchwork")
library("cyCombine")
library("cyDefine") # Available at https://github.com/biosurf/cyDefine
library("scennep")
library("spaceRATScaffolds") # Available at https://github.com/shdam/spaceRATScaffolds
seed <- 942
mc.cores <- 20
```


## The CITE-seq dataset

We use the Seurat PBMC atlas by [Hao et al., 2021](https://www.cell.com/cell/fulltext/S0092-8674(21)00583-3)

### ADT


```{r load_data}
if (file.exists("data/01_citeseq_seurat_adt.RDS")) {
  adt <- readRDS("data/01_citeseq_seurat_adt.RDS")
} else {
  seu_pbmc <- LoadH5Seurat("/home/projects/dp_immunoth/people/s153398/cytograted/data/aml/scrna-seq/pbmc_multimodal.h5seurat",
                            assays = c("ADT"),
                            reductions = FALSE,
                            graphs = FALSE,
                            images = FALSE)
  
  seu_pbmc$batch <- seu_pbmc$orig.ident
  seu_pbmc$sample <- "ADT"
  suppressMessages({ # Hide per-cluster correction info
  seu_pbmc <- seu_pbmc |> 
    cyCombine(
      norm_method = "clr",
      markers = rownames(seu_pbmc), 
      xdim = 4, ydim = 3, 
      cluster_method = "kohonen", 
      layer = "data",
      mc.cores = 1
    )
  })
  
  
  adt <- seu_pbmc |> 
    GetAssayData("cyCombine", "data") |>
    t() |> 
    tibble::as_tibble() |> 
    mutate(
      sample = "ADT",
      batch = "ADT",
      cell_id = colnames(seu_pbmc),
      celltype = seu_pbmc$celltype.l2,
      celltype = dplyr::case_when(
    startsWith(celltype, "B") ~ "B cells",
    endsWith(celltype, "Mono") ~ "Monocytes",
    startsWith(celltype, "CD4") ~ "CD4pos",
    celltype == "Treg" ~ "CD4pos",
    startsWith(celltype, "CD8") ~ "CD8pos",
    startsWith(celltype, "NK") ~ "NK cells",
    celltype == "gdT" ~ "TCRgd",
    celltype == "pDC" ~ "pDCs",
    startsWith(celltype, "cDC") ~ "cDCs",
    celltype == "ASDC" ~ "cDCs",
    celltype == "TCRgd" ~ celltype,
    # celltype == "Platelets" ~ celltype,
    TRUE ~ "Other"
  ))
}

```

### Map CITE-seq markers to gene names

```{r map_markers}
if (!file.exists("data/01_citeseq_seurat_adt.RDS")) {
  # Download marker info
  if (!file.exists("ab_info.xlsx")) system("wget https://ars.els-cdn.com/content/image/1-s2.0-S0092867421005833-mmc1.xlsx -O ab_info.xlsx")
  
  # Load marker mapper
  gene_marker <- readxl::read_excel("ab_info.xlsx")
  
  # gene_marker <- 
  gene_marker$`Ensembl Gene Id` <- gene_marker$`Ensembl Gene Id` |> stringr::str_remove_all(".*MUS.* ")
  gene_marker$`#protein` <- gsub("_", "-", gene_marker$`#protein`)
  # Select relevant genes-marker pairs
  gm <- sapply(colnames(adt), function(m) gene_marker$`Ensembl Gene Id`[stringr::str_detect(gene_marker$`#protein`, m)][1])
  
  # Manually assign gene IDs for remaining markers
  gm[is.na(gm)]
  gm[c("CD11a/CD18", "CD18") ] <- c("ENSG00000005844", "ENSG00000160255")
  gm <- gsub("ENSMUSG\\d+, ", "", gm)
  gm <- gsub("NA", NA, gm)

  gm <- gm[!is.na(gm)]
  
  colnames(adt)[which(colnames(adt) %in% names(gm))] <- gm[colnames(adt)[which(colnames(adt) %in% names(gm))]]
  c <- colnames(adt)
  duplicated_genes <- unique(c[duplicated(c) & !is.na(c)])
  duplicated_location <- which(c %in% duplicated_genes)
  
  dup_pbmc <- lapply(setNames(duplicated_genes, duplicated_genes), function(gene) {
    dup_max <- apply(adt[, gm[gm == gene]], 1, max)
    return(dup_max)
  })
  adt[, duplicated_location] <- NULL
  
  adt <- cbind(adt, dup_pbmc)
  
  adt <- adt |> 
    dplyr::select(dplyr::any_of(non_markers), dplyr::starts_with("ENSG"))
  
  saveRDS(adt, file = "data/01_citeseq_seurat_adt.RDS")
}
```

### RNA

```{r rna}
if (file.exists("data/01_citeseq_seurat_rna.RDS")) {
  seu_rna <- readRDS("data/01_citeseq_seurat_rna.RDS")
} else {

# RNAseq
# devtools::load_all("../spaceRATScaffolds")
seu_rna <- LoadH5Seurat("/home/projects/dp_immunoth/people/s153398/cytograted/data/aml/scrna-seq/pbmc_multimodal.h5seurat",
                    assays = c("SCT"),
                    reductions = FALSE,
                    graphs = FALSE,
                    images = FALSE)


rna <- GetAssayData(seu_rna, layer = "data")
counts <- GetAssayData(seu_rna, layer = "counts")
md <- seu_rna[[]]

data(gene_id_converter_hs, package = "spaceRATScaffolds")

convert_gene2id <- function(genes) {
  gene_id_converter_hs$ensembl_gene[match(genes, gene_id_converter_hs$hgnc_symbol)]
}

rownames(rna) <- convert_gene2id(rownames(rna))
rna <- rna[!is.na(rownames(rna)),]
rownames(counts) <- convert_gene2id(rownames(counts))
counts <- counts[!is.na(rownames(counts)),]

variable_feats <- convert_gene2id(VariableFeatures(seu_rna))
variable_feats <- variable_feats[!is.na(variable_feats)]

seu_rna <- CreateSeuratObject(counts = counts, data = rna, assay = "SCT")
VariableFeatures(seu_rna) <- variable_feats
seu_rna <- AddMetaData(seu_rna, md)


celltype <- seu_rna$celltype.l2
seu_rna$celltype <- dplyr::case_when(
  startsWith(celltype, "B") ~ "B cells",
  endsWith(celltype, "Mono") ~ "Monocytes",
  startsWith(celltype, "CD4") ~ "CD4pos",
  startsWith(celltype, "CD8") ~ "CD8pos",
  celltype == "NK" ~ "NK cells",
  celltype == "pDC" ~ "pDCs",
  startsWith(celltype, "cDC") ~ "cDCs",
  celltype == "TCRgd" ~ celltype,
  celltype == "Platelets" ~ celltype,
  TRUE ~ "Other"
) |> as.character()

saveRDS(seu_rna, "data/01_citeseq_seurat_rna.RDS")
}
```

## Pseudobulk neighbors of RNA-part

```{r scennep, warning=FALSE}
# Common markers
markers <- intersect(colnames(adt), rownames(seu_rna))

if (file.exists("data/citeseq_scennep_rna.RDS")) {
  rna <- readRDS("data/citeseq_scennep_rna.RDS")
} else {

rna <- scennep(
  seu_rna,
  mc.cores = 30,
  markers = markers,
  flavor = "SCT",
  distance = "cosine",
  pb = interactive(), 
  return_S4 = FALSE,
)

saveRDS(seu_rna, "data/citeseq_scennep_rna.RDS")
}
```

## Merge datasets

```{r merge}
#GetAssayData(rna, "scennep","data") |> 
# rna[["SCT"]]$data[markers,] |>
rna_df <- rna |> 
  t() |>
  tibble::as_tibble() |> 
  dplyr::mutate(
    batch = "scRNA-seq",
    sample = "scRNA-seq",
    cell_id = colnames(rna),
    celltype = rna$celltype
) |> 
  dplyr::filter(cell_id %in% adt$cell_id)

# Merge datasets
set.seed(seed)
uncorrected <- rna_df |> 
  dplyr::bind_rows(adt) |> 
  dplyr::select(dplyr::any_of(c(markers, non_markers))) |> 
  dplyr::mutate(id = dplyr::row_number())

# saveRDS(uncorrected, "results/04_citeseq_adt_rna_uncorrected_scennep.RDS")
```

## Uncorrected

```{r uncorrected}
set.seed(seed)
uncor_batch <- plot_umap(
  uncorrected,
  down_sample = T,
  sample_n = 5000,
  markers = markers,
  title = "Uncorrected",
  ref_col = "batch")

# saveRDS(uncor_batch, "results/04_citeseq_adt_rna_uncor.RDS")
# uncor_cytof + uncor_rnaseq +
#   patchwork::plot_layout(guides = "collect")
uncor_batch
```

## Batch correct
```{r correct}

system.time({
  corrected <- cyCombine(
    uncorrected,
    markers = markers,
    norm_method = "scale",
    cluster_method = "flowsom",
    distf = "cosine",
    seed = seed,
    xdim = c(1,1,5),
    ydim = c(1,5),
    mc.cores = 1
  )
})
# saveRDS(corrected, "results/04_citeseq_adt_rna_corrected.RDS")
```

## Plot results - Cells

```{r plot_cells}
colors <- get_distinct_colors(sort(unique(corrected$celltype)))
set.seed(seed)

plot_cell <- plot_umap(
  corrected[corrected$batch == "scRNA-seq",],
  corrected[corrected$batch == "ADT",],
  down_sample = T,
  sample_n = 5000,
  markers = markers,
  title = c("RNA", "ADT"),
  ref_col = "celltype",
  colors = colors,
  query_col = "celltype")

# saveRDS(plot_cell, "results/04_citeseq_adt_rna_cell.RDS")
# ggplot2::ggsave(plot_cell, filename = "figs/04_citeseq_adt_rna_cell.png")
plot_cell

```

## Plot results - Batch

```{r plot_batch}
set.seed(seed)
umap_batch <- plot_umap(
  corrected,
  down_sample = T, 
  sample_n = 5000,
  markers = markers,
  title = "CITEseq - ADT + RNA",
  ref_col = "batch")

# ggplot2::ggsave(umap_batch, filename = "figs/04_citeseq_adt_rna_batch.png")
# saveRDS(umap_batch, "results/04_citeseq_adt_rna_batch.RDS")
umap_batch
```


## Measure performance

```{r}
if (! exists("uncorrected")) uncorrected <- readRDS("results/04_citeseq_adt_rna_uncorrected_SCT.RDS")
if (! exists("corrected")) corrected <- readRDS("results/04_citeseq_adt_rna_corrected_SCT.RDS")
if (! exists("markers")) markers <- cyCombine::get_markers(uncorrected)
```


```{r performance}


corrected$label <- create_som(corrected, ydim = 5, xdim = 5, seed = seed, markers = markers, cluster_method = "flowsom", distf = "cosine")
uncorrected$label <- corrected$label

system.time({
  lu <- cyCombine::compute_cilisi(df = uncorrected, markers = get_markers(uncorrected), label_cols = "batch", split_by = "label", return_mean = F)
})
message("CiLISI uncorrected: ", mean(unlist(lu)))

system.time({
  l <- cyCombine::compute_cilisi(df = corrected, markers = get_markers(corrected), label_cols = "batch", split_by = "label", return_mean = F)
})
message("CiLISI corrected: ", mean(unlist(l)))

mad <- cyCombine::evaluate_mad(uncorrected, corrected, cell_col = "label", markers = markers)
emd <- cyCombine::evaluate_emd(uncorrected, corrected, cell_col = "label", markers = markers)
# saveRDS(list("emd" = emd, "mad" = mad, "CiLISI_uc" = lu, "CiLISI_c" = l), "results/04_citeseq_adt_rna_performance_SCT.RDS")

```

## Measure distance


```{r distance}
if ("cell" %in% colnames(uncorrected)) { # Fix legacy names
  uncorrected <- rename(uncorrected, celltype = cell)
  corrected <- rename(corrected, celltype = cell)
}
APPLY <- cyCombine:::set_apply(mc.cores, pb = TRUE)
markers <- get_markers(uncorrected)
colors <- get_distinct_colors(sort(unique(corrected$celltype)))
distances_uncorrected <- uncorrected |> 
  group_by(cell_id) |> 
  dplyr::group_split() |> 
  APPLY(function(df) {
    proxy::dist(as.matrix(df[, markers]), method = "cosine")
  }, mc.cores = mc.cores)
distances_uncorrected <- do.call(rbind, distances_uncorrected)
distances_uncorrected_df <- data.frame(
  "distance" = distances_uncorrected,
  "type" = "Uncorrected"
)


distances <- corrected |> 
  group_by(cell_id) |> 
  dplyr::group_split() |> 
  APPLY(function(df) {
    proxy::dist(as.matrix(df[, markers]), method = "cosine")
  }, mc.cores = mc.cores)
# distances <- do.call(rbind, distances)
distances_df <- data.frame(
  "distance" = do.call(rbind, distances),
  "type" = "Corrected"
)
# distances_df_celltype <- distances |> rename(type = cell) |> 
  # select(distance, type)

distances_celltypes <- corrected |> 
  dplyr::filter(batch == "ADT") |> 
  group_by(celltype) |> 
  summarize(across(all_of(markers), mean, rm.na=TRUE)) |> 
  select(all_of(markers)) |> 
  proxy::dist(method = "cosine")
#distances_celltypes <- do.call(rbind, distances_celltypes)
distances_celltypes_df <- data.frame(
  "distance" = distances_celltypes,
  "type" = "Celltype"
)

set.seed(seed)
distances_celltypes_pop <- corrected |> 
  dplyr::filter(batch == "ADT") |> 
  dplyr::group_by(celltype) |> 
  dplyr::slice_sample(n = 1000) |> 
  dplyr::group_split(.keep = TRUE) |>
  APPLY(function(df) {
    d <- proxy::dist(as.matrix(df[, markers]), method = "cosine") |> 
      as.numeric()
    d <- dplyr::tibble(distance = d, type = df$celltype[1])
    return(d)
  }, mc.cores = mc.cores)
  # summarize(across(any_of(markers), mean, rm.na=TRUE)) |> 
  # select(markers) |> 
  # dist()
#distances_celltypes <- do.call(rbind, distances_celltypes)
distances_celltypes_pop <- do.call(rbind, distances_celltypes_pop)
# distances_celltypes_pop_df <- data.frame(
#   "distance" = distances_celltypes_pop,
#   "type" = "Population"
# )

combined_distances <- rbind(distances_uncorrected_df, distances_df, distances_celltypes_pop)
combined_distances$type <- factor(combined_distances$type, levels = c(
  "Corrected", "Uncorrected",
  "B cells", "CD4pos", "CD8pos", "Monocytes", "NK cells", "cDCs", "pDCs"
))

saveRDS(combined_distances, "results/04_citeseq_adt_rna_distances_SCT.RDS")
# combined_distances <- readRDS("results/04_citeseq_adt_rna_distances.RDS")
combined_distances$size <- as.numeric(case_when(combined_distances$type %in% c("Corrected", "Uncorrected") ~ 0.6, TRUE ~ .5))
plot_distance <- ggplot(combined_distances, aes(x = distance, color = type)) +
  geom_density(linewidth = .8) +
  labs(title = "Cell-by-cell distances",
       x = "Distance",
       y = "Density", color = "Type") +
  theme_bw() + 
  scale_color_manual(values = c(colors, "Corrected" = "black", "Uncorrected" = "gray"))

# ggplot2::ggsave(plot_distance, filename = "figs/04_citeseq_adt_rna_distance_SCT.png")
plot_distance
# ggsave(plot_distance, filename = "figs/04_citeseq_adt_rna_distance_SCT.pdf", height = 10, width = 20, units = "cm")
```

```{r plots, include=FALSE}
plots <- plot_cell / plot_distance +
  plot_layout(tag_level = 'new') + 
  plot_annotation(tag_levels = "A") +
  theme(plot.tag = element_text(size = 14))
# ggsave(plots, filename = "figs/04_citeseq_adt_rna_cell_distance_SCT.png", height = 20, width = 20, units = "cm")
# ggsave(plots, filename = "figs/04_citeseq_adt_rna_cell_distance_SCT.pdf", height = 20, width = 20, units = "cm")
```
