---
title: "Adapting the PBMC Reference"
author: "Søren H Dam"
date: "Last updated: `r format(Sys.Date(), '%Y.%m.%d')`"
output: 
  html_document:
    toc: true
    toc_float: true
    highlight: kate
    df_print: paged
---

<style type="text/css">

h1.title {
  color: #004d66;
}
h4.author {
  font-style: italic;
  font-size: 18px;
  color: #008cba;
}
h4.date {
  font-style: italic;
  font-size: 16px;
  color: #008cba;
}
h1 { /* Header 1 */
  color: #004d66;
  font-size: 28px;
}
h2 { /* Header 2 */
  color: #004d66;
  font-size: 22px;
}
h3 { /* Header 3 */
  color: #004d66;
  font-size: 16px;
}
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Adapting a reference to the query markers

This vignette demonstrates the process of adapting a reference dataset, specifically a reference of Peripheral Blood Mononuclear Cells (PBMCs), to match the marker panels of different query datasets.
Reference adaptation is sometimes necessary because it is not always possible to achieve the same level of annotation granularity with a smaller set of markers than what is possible in a reference set with many markers.
The adaptation involves merging populations that cyDefine fails to distinguish between.

This vignette demonstrates the adaptation process using a CyTOF (Z2YR) and Spectral Flow Cytometry (OMIP69) panel as well as the small OMIP23 panel. 


## Dependencies

```{r dependencies}
library(dplyr)
library(ggplot2)
library(patchwork)
library(cyCombine)
library(cyDefine)

```

## The panels

```{r, panels}
# Set up seed and number of cells
seed <- 430
n_cells <- 30000


# HIMC (FR-FCM-Z2YR) panel
himc_panel <- c("CD45", "CD57", "HLA-DR", "CCR6", "CD19", "IgG", "CD4", "IgD", "CD20", "IgA", "CD56", "CD86", "ICOS", "TCRgd", "CD45RA", "CD123", "CD27", "CXCR3", "NKG2A", "anti", "CD11c", "CD14", "CD26", "CD8a", "CD33", "CD161", "CD127", "CCR10", "CCR7", "CCR9", "CD25", "CD3", "CXCR5", "CD38", "a4b7", "PD-1", "CD62L", "CLA", "CD16")

# OMIP-023 markers
omip23 <- c("CD45", "CD3", "CD8", "CD4", "CD25", "CD127", "CD19", "CD38", "HLA-DR", "CD16", "CD56", "CD14", "CD69")

# OMIP-069 markers
omip69 <- c("CD45","CD3","CD4","CD8","CD25","TCRgd","CD14","CD16","CD11c","CD19","CD20","CD24","CD39","IgD","IgG","IgM","CD141","CD1c","CD123","CD2","CD56","CCR7","CD27","CD28","CD45RA","CD95","CD127","CD337","CCR6","CCR5","CXCR5","CXCR3","HLADR","CD38","CD57","PD-1","CD159a","CD159c","CD314")

# POISED markers
poised_panel <- c("CD16", "IFNg", "CD69", "CXCR3", "IL.17", "LAP",
             "CD27", "CD40L", "PD1", "CD123", "CD45RA", "CD28",
             "GPR15", "HLA.DR", "CD33", "CD14", "CD127", "CD86",
             "TCRgd", "CD19", "CD49b", "IL.4", "CD4", "CD8",
             "CD38", "LAG3", "OX40", "CD20", "CCR4", "IL9",
             "CD3", "CD11c", "CCR7", "CD57", "IL.10", "integrin",
             "CD25", "CD56", "CLA")
```

## The Seurat PBMC reference

```{r, plot_ref}
# Get and filter markers from Seurat reference
pbmc_reference <- get_reference("pbmc", path = "data", store = TRUE)
ref_markers <- get_markers(pbmc_reference)
ref_markers <- ref_markers[!startsWith(ref_markers, "cell") & !startsWith(ref_markers, "Rat")]

# Subset reference
reference <- pbmc_reference |> 
    dplyr::filter(
      !celltype.l2 %in% c("Doublet", "Platelet", "Eryth", "HSPC")
  )
# Set up color palette for cell types
set.seed(seed)
colors <- reference |> 
  pull(celltype.l2) |> 
  unique() |> sort() |> 
  get_distinct_colors()

# UMAP plot for Seurat reference
umap_ref <- plot_umap(
  reference,
  return_data = T,
  sample_n = n_cells,
  markers = ref_markers,
  add_centroids = "text",
  title = "Universal PBMC reference",
  col = "celltype.l2")

umap_ref$plot
```
## OMIP23 - Adapt reference

### Map markers

```{r, mapped_dummy_omip23}
# Create a dummy query data frame based on OMIP23 markers
dummy_query <- data.frame(matrix(0, nrow = 1, ncol = length(omip23)))
colnames(dummy_query) <- omip23

# Map reference markers to OMIP23 panel
mapped_dummy_omip23 <- map_marker_names(ref_markers = ref_markers,
  using_pbmc = TRUE, query = dummy_query,
  query_markers = omip23
)

```

### Reference on query markers

```{r, plot_ref_omip23}
# Generate UMAP plot using OMIP23 markers
umap_ref_omip23 <- plot_umap(
  umap_ref$data,
  down_sample = F,
  return_data = T,
  markers = colnames(mapped_dummy_omip23),
  add_centroids = "text",
  title = "OMIP-023 marker subset",
  col = "celltype.l2")

umap_ref_omip23$plot
```


### Adapt OMIP23 to reference

```{r, adapted_reference_omip23}
# Adapt the Seurat reference to match OMIP23 markers
adapted_reference_omip23 <- adapt_reference(
  reference = reference,
  markers = colnames(mapped_dummy_omip23),
  mtry = floor(ncol(mapped_dummy_omip23)/2),
  num.threads = 4,
  # exclude_celltypes = c("Doublet", "Platelet", "Eryth", "HSPC"),
  using_pbmc = TRUE,
  seed = seed
)
```

```{r adapted_celltypes_omip23}
# List names to add custom colors (optional)
table(adapted_reference_omip23$celltype)
```

```{r, plot_adapted_omip23, out.width="100%"}

colors <- cyDefine:::expand_colors(adapted_celltypes = adapted_reference_omip23$celltype, original_celltypes = reference$celltype.l2, colors = colors)
# Join adapted cell types and plot the UMAP
plot_adapted_omip23 <- umap_ref_omip23$data |> 
  dplyr::left_join(adapted_reference_omip23[, c("cell_id", "celltype", "celltype_original")], by = c("cell_id")) |> 
  plot_embedding(
    add_centroids = "text",
    title = "Adapted reference for OMIP-023",
    highlight_labels = T,
    colors = colors,
    col = "celltype")
umap_ref_omip23$plot + plot_adapted_omip23
```

```{r diagram_omip23}
# Generate a diagram visualizing the adapted reference structure
fontcolor_nodes <- c(
  "NK" = "white", 
  "NK Proliferating / NK" = "white"
)

plot_diagram(adapted_reference_omip23, colors = colors, fontcolor_nodes = fontcolor_nodes)
```



## OMIP69 - Adapt reference

### Map markers
```{r, mapped_dummy_omip69}
# Create a dummy query data frame based on OMIP69 markers
dummy_query <- data.frame(matrix(0, nrow = 1, ncol = length(omip69)))
colnames(dummy_query) <- omip69

# Map reference markers to OMIP69 panel
mapped_dummy_omip69 <- map_marker_names(
  ref_markers = ref_markers,
  using_pbmc = TRUE, query = dummy_query,
  query_markers = omip69
)
```

### Reference on query markers

```{r, plot_ref_omip69}
# Generate UMAP plot using OMIP69 markers
umap_ref_omip69 <- plot_umap(
  umap_ref$data,
  return_data = T,
  down_sample = F,
  add_centroids = "text",
  title = "OMIP-069 marker subset",
  markers = colnames(mapped_dummy_omip69),
  col = "celltype.l2")

umap_ref_omip69$plot
```

### Adapt reference
```{r, adapted_reference_omip69}
# Adapt the Seurat reference to match OMIP69 markers
adapted_reference_omip69 <- adapt_reference(
  reference = reference,
  markers = colnames(mapped_dummy_omip69),
  mtry = floor(ncol(mapped_dummy_omip69)/2),
  num.threads = 4,
  # exclude_celltypes = c("Doublet", "Platelet", "Eryth", "HSPC"),
  using_pbmc = TRUE,
  seed = seed
)
```

```{r, plot_adapted_omip69, out.width="100%"}
# Adjust color mapping for specific cell types
colors <- cyDefine:::expand_colors(adapted_celltypes = adapted_reference_omip69$celltype, original_celltypes = reference$celltype.l2, colors = colors)
# Join adapted cell types and plot the UMAP
plot_adapted_omip69 <- umap_ref_omip69$data |> 
  dplyr::left_join(adapted_reference_omip69[, c("cell_id", "celltype", "celltype_original")], by = c("cell_id")) |> 
  plot_embedding(
    add_centroids = "text",
    highlight_labels = T,
    title = "Adapted reference for OMIP-069",
    colors = colors,
    col = "celltype")
umap_ref_omip69$plot + plot_adapted_omip69
```




```{r diagram_omip69}
# Generate a diagram visualizing which cell types had to be merged
fontcolor_nodes <- c(
  "NK" = "white", 
  "NK Proliferating / NK" = "white"
)

plot_diagram(adapted_reference_omip69, colors = colors, fontcolor_nodes = fontcolor_nodes)
```

## Figure 2

```{r full_plot_small, out.width="100%"}
layout <- "
##BBCC
AABBCC
AADDEE
##DDEE
"

full_plot_small <- umap_ref$plot + umap_ref_omip23$plot + plot_adapted_omip23 + umap_ref_omip69$plot + plot_adapted_omip69 +
  plot_layout(design = layout) +
  patchwork::plot_annotation(tag_levels = "a") &
  theme(text = element_text(size = 12))


# full_plot
# ggsave(full_plot_small, filename = "../figs/adapt_reference_omip23_omip69.pdf", units = "cm", height = 25, width = 35)
full_plot_small
```


## Z2YR - Adapt reference

### Map markers
```{r, mapped_dummy_himc}
# Create a dummy query data frame based on HIMC markers
dummy_query <- data.frame(matrix(0, nrow = 1, ncol = length(himc_panel)))
colnames(dummy_query) <- himc_panel

# Map reference markers to HIMC panel
mapped_dummy_himc <- map_marker_names(ref_markers = ref_markers,
  using_pbmc = TRUE, query = dummy_query,
  query_markers = himc_panel
)

```

### Reference on query markers

```{r, plot_ref_himc}

# Generate UMAP plot using HIMC markers
umap_ref_himc <- plot_umap(
  umap_ref$data,
  down_sample = F,
  return_data = T,
  add_centroids = "text",
  title = "Z2YR marker subset (CyTOF)",
  colors = colors,
  markers = colnames(mapped_dummy_himc),
  col = "celltype.l2")

umap_ref_himc$plot
```


### Adapt Z2YR to reference

```{r, adapted_reference_himc}
# Adapt the Seurat reference to match HIMC markers
adapted_reference_himc <- adapt_reference(
  reference = reference,
  markers = colnames(mapped_dummy_himc),
  mtry = floor(ncol(mapped_dummy_himc)/2),
  num.threads = 4,
  # exclude_celltypes = c("Doublet", "Platelet", "Eryth", "HSPC"),
  using_pbmc = TRUE,
  seed = seed
)
```

```{r adapted_celltypes}
# List names to add custom colors (optional)
table(adapted_reference_himc$celltype)
```

```{r, plot_adapted_himc, out.width="100%"}
# Adjust color mapping for specific cell types
# colors["CD8 TCM / CD8 Naive"] <- colors["CD8 Naive"]
# colors["cDC"] <- colors["cDC1"]
colors <- cyDefine:::expand_colors(adapted_celltypes = adapted_reference_himc$celltype, original_celltypes = reference$celltype.l2, colors = colors)
# Join adapted cell types and plot the UMAP
plot_adapted_himc <- umap_ref_himc$data |> 
  dplyr::left_join(adapted_reference_himc[, c("cell_id", "celltype", "celltype_original")], by = c("cell_id")) |> 
  plot_embedding(
    add_centroids = "text",
    title = "Adapted reference for Z2YR (CyTOF)",
    highlight_labels = T,
    colors = colors,
    col = "celltype")
umap_ref_himc$plot + plot_adapted_himc
```

```{r diagram_himc}
# Generate a diagram visualizing the adapted reference structure
fontcolor_nodes <- c(
  "NK" = "white", 
  "NK Proliferating / NK" = "white"
)

plot_diagram(adapted_reference_himc, colors = colors, fontcolor_nodes = fontcolor_nodes)
```


## POISED - Adapt reference

### Map markers
```{r, mapped_dummy_poised}
# Create a dummy query data frame based on OMIP23 markers
dummy_query <- data.frame(matrix(0, nrow = 1, ncol = length(poised_panel)))
colnames(dummy_query) <- poised_panel

# Map reference markers to OMIP23 panel
mapped_dummy_poised <- map_marker_names(ref_markers = ref_markers,
  using_pbmc = TRUE, query = dummy_query,
  query_markers = poised_panel, map_specific_from = c("integrin"), map_specific_to = c("Integrin-7")
)

```

### Reference on query markers

```{r, plot_ref_poised}
# Generate UMAP plot using POISED markers
umap_ref_poised <- plot_umap(
  umap_ref$data,
  down_sample = F,
  return_data = T,
  markers = colnames(mapped_dummy_poised),
  add_centroids = "text",
  title = "POISED marker subset",
  col = "celltype.l2")

umap_ref_poised$plot
```


### Adapt POISED to reference

```{r, adapted_reference_poised}
# Adapt the Seurat reference to match POISED markers
adapted_reference_poised <- adapt_reference(
  reference = reference,
  markers = colnames(mapped_dummy_poised),
  mtry = floor(ncol(mapped_dummy_poised)/2),
  num.threads = 4,
  # exclude_celltypes = c("Doublet", "Platelet", "Eryth", "HSPC"),
  using_pbmc = TRUE,
  seed = seed
)
```

```{r adapted_celltypes_poised}
# List names to add custom colors (optional)
table(adapted_reference_poised$celltype)
```

```{r, plot_adapted_poised, out.width="100%"}

colors <- cyDefine:::expand_colors(adapted_celltypes = adapted_reference_poised$celltype, original_celltypes = reference$celltype.l2, colors = colors)
# Join adapted cell types and plot the UMAP
plot_adapted_poised <- umap_ref_poised$data |> 
  dplyr::left_join(adapted_reference_poised[, c("cell_id", "celltype", "celltype_original")], by = c("cell_id")) |> 
  plot_embedding(
    add_centroids = "text",
    title = "Adapted reference for POISED",
    highlight_labels = T,
    colors = colors,
    col = "celltype")
umap_ref_poised$plot + plot_adapted_poised
```

```{r diagram_poised}
# Generate a diagram visualizing the adapted reference structure
fontcolor_nodes <- c(
  "NK" = "white", 
  "NK Proliferating / NK" = "white"
)

plot_diagram(adapted_reference_poised, colors = colors, fontcolor_nodes = fontcolor_nodes)
```






