---
title: "Phenotype transfer between CyTOF and scRNA-seq"
author: "Søren H Dam"
date: "Last updated on `r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    highlight: kate
    df_print: paged
---

<style type="text/css">

h1.title {
  color: #004d66;
}
h4.author {
  font-style: italic;
  font-size: 18px;
  color: #008cba;
}
h4.date {
  font-style: italic;
  font-size: 16px;
  color: #008cba;
}
h1 { /* Header 1 */
  color: #004d66;
  font-size: 28px;
}
h2 { /* Header 2 */
  color: #004d66;
  font-size: 22px;
}
h3 { /* Header 3 */
  color: #004d66;
  font-size: 16px;
}
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



This vignette demonstrates integrating a split-sample scRNA-seq and CyTOF experiment ([Su et al.](https://www.nature.com/articles/s41597-024-03399-6)). The purpose is to transfer phenotypes from the CyTOF annotation to the scRNA-seq data, thus automating its annotation. The hope is that this process will identify cell populations otherwise only found in the CyTOF dataset, although at the expense of not finding populations only found by manually annotating the scRNA-seq data.
To assimilate scRNA-seq data to CyTOF, we use our own implementation of Single-cell nearest-neighbor pseudobulking ([scennep](https://github.com/shdam/scennep)).

## Dependencies

```{r dependencies}
library("Seurat")
library("anndata")
library("dplyr")
library("ggplot2")
library("patchwork")
library("cyCombine")
library("cyDefine") # Available at https://github.com/biosurf/cyDefine
library("scennep") # Available at https://github.com/shdam/scennep
seed <- 447
mc.cores <- 20
```

## Download data 

We got the download instructions from the article's [Data availability section](https://www.nature.com/articles/s41597-024-03399-6#Sec7). CyTOF is available on FlowRepository ([FR-FCM-Z6ZN](https://flowrepository.org/id/FR-FCM-Z6ZN)) and scRNA-seq data is available on SRA ([SRR23497303](https://trace.ncbi.nlm.nih.gov/Traces/?run=SRR23497303)). Getting the scRNA-seq metadata took some looking around, but direct download links are presented in the code chunks.

### Get CyTOF data (FR-FCM-Z6ZN)

FlowRepositoryR might be depricated, so manual download can be required.

```{r get-cytof, eval = FALSE}
if (!require("FlowRepositoryR")) {
  remotes::install_github("jspidlen/FlowRepositoryR")
}


library(FlowRepositoryR)

x <- flowRep.get("FR-FCM-Z6ZN")

x <- download(x)

```

### Get scRNA-seq data (SRR23497303)

Run in bash.

```{bash get-rna, eval = FALSE}
cd ../../../data/pbmc/paired_cytometry_scrna/rnaseq

# SRA Normalized 
#wget https://sra-pub-run-odp.s3.amazonaws.com/sra/SRR23497303/SRR23497303
# Raw bam file (used)
wget https://sra-pub-src-1.s3.amazonaws.com/SRR23497303/possorted_genome_bam.bam.1
mv possorted_genome_bam.bam.1 possorted_genome.bam

# Metadata
wget https://ftp.ncbi.nlm.nih.gov/geo/samples/GSM7048nnn/GSM7048570/suppl/GSM7048570%5Fpbmc%5Fseq.h5ad.gz
gunzip rnaseq/GSM7048570_pbmc_seq.h5ad.gz
rm rnaseq/GSM7048570_pbmc_seq.h5ad.gz
```


## Preprocessing

Before I found the preprocessed data from the article, I manually rerun the preprocessing steps. In the end, I used my own preprocessing, but used their filtering. There is commented out code for reusing their expression data, if you are redoing the analysis and want to keep it simple.

### scRNA-seq

First, I convert the bam file to fastq in order to reanalyze with Cell Ranger.

```{bash cellranger, eval = FALSE}
module load cellranger/9.0.1
module load bamtofastq/1.3.0

# Get reference
wget "https://cf.10xgenomics.com/supp/cell-exp/refdata-gex-GRCh38-2024-A.tar.gz"
tar -zxvf refdata-gex-GRCh38-2024-A.tar.gz

# Bam to fastq 
bamtofastq --nthreads 38 possorted_genome.bam fastq
mv fastq/ES-PBMC-2_HiSeq_MissingLibrary_1_HJK5NBCX2/* fastq/.
rm -rf fastq/ES-PBMC-2_HiSeq_MissingLibrary_1_HJK5NBCX

# Cell Ranger
cellranger count --transcriptome refdata-gex-GRCh38-2024-A --fastqs fastq --localcores 39 --localmem 110 --output-dir cr_out --id cr_paired_rna --create-bam=false
```

Then, I load the output into Seurat for preprocessing.

```{r seurat}
seu <- Seurat::Read10X("../../../data/pbmc/paired_cytometry_scrna/rnaseq/cr_out/outs/filtered_feature_bc_matrix")
seu <- CreateSeuratObject(counts = seu, project = "paired_rna", min.cells = 3, min.features = 200)
seu[["percent.mt"]] <- PercentageFeatureSet(seu, pattern = "^MT-")

# VlnPlot(seu, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

(Commented out) I manually set the processing thresholds as follows:

```{r subset}
# seu <- subset(seu, subset = nFeature_RNA > 250 & nFeature_RNA < 2800 & percent.mt < 5)
# VlnPlot(seu, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

I reused the filtering and annotation from Su et al.

```{r extract_metadata}
# Load anndata
adata <- read_h5ad("../../../data/pbmc/paired_cytometry_scrna/rnaseq/GSM7048570_pbmc_seq.h5ad")
# 
# # Access the data
# count_matrix <- adata$X
# seu <- Seurat::CreateSeuratObject(t(adata$raw$X))
# seu <- seu[, adata$raw$obs_names %in% rownames(metadata)]
# count_matrix <- count_matrix[adata$raw$obs_names %in% rownames(metadata), ]
metadata <- adata$obs
# gene_info <- adata$var

# seu <- subset(seu, subset = nFeature_RNA > 250 & nFeature_RNA < 2800 & percent.mt < 10)
# seu <- subset(seu, subset = nFeature_RNA > 250 & (nFeature_RNA < quantile(nFeature_RNA, .95)) & percent.mt < 10)
seu <- seu[, colnames(seu) %in% rownames(metadata)]

# rna <- tibble::as_tibble(count_matrix) |> 
#   mutate(cell_id = rownames(count_matrix),
#          celltype = metadata$celltype) 

seu <- AddMetaData(seu, metadata)
# seu$in_orig <- if_else(colnames(seu) %in% rownames(metadata), true = "In", false = "Not in")
```

#### Visualizing scRNA-seq

```{r umap}
set.seed(seed)
# Normalize and PCA
seu <- NormalizeData(seu)
seu <- FindVariableFeatures(seu, selection.method = "vst", nfeatures = 3000)
seu <- ScaleData(seu)
seu <- RunPCA(seu)


# Cluster
seu <- FindNeighbors(seu, dims = 1:10)
seu <- FindClusters(seu, algorithm = 4, random.seed = seed)

# UMAP
seu <- RunUMAP(seu, dims = 1:10, seed.use = seed)
DimPlot(seu, reduction = "umap", group.by = "celltype")

# DimPlot(seu, reduction = "umap", group.by = "in_orig")
```


### CyTOF

Load FCS files into R and clean marker names.

```{r prep_cytof}
data_dir <- "../../../data/pbmc/paired_cytometry_scrna/FR-FCM-Z6ZN/"
cytof_fs <- cyCombine::compile_fcs(data_dir, pattern = "analyze_normalized")
cytof_fs[[1]] |> flowCore::parameters() |> Biobase::pData() |> 
  dplyr::filter(stringr::str_detect(desc, "_"),
         stringr::str_detect(desc, "BC", negate = T)) |> 
  pull("desc") 
cytof <- prepare_data(flowset = cytof_fs, clean_colnames = F, down_sample = F, transform = T)


colnames(cytof) <- colnames(cytof) |> 
  stringr::str_remove_all("^\\d+[A-Za-z]+_?")
cytof <- cytof[, stringr::str_detect(colnames(cytof), "^(CD|Fox|Tbet|id|sample|IgD|HLA)")]

markers <- cyCombine::get_markers(cytof)
```

#### Annotation

I could not extract the annotation used in Su et al., so I manually annotated using more or less the same approach as they do.

```{r cy_annotation}

cytof$label <- create_som(
  cytof, seed = 334, cluster_method = "flowsom", xdim = 6, ydim = 6, nClus = 18)

umap_cytof <- cytof |> 
  plot_umap(markers = markers, col = "label", seed = 334, return_data = TRUE, down_sample = T, sample_n = 30000, title = "CyTOF UMAP")
umap_cytof$plot
# plot_embedding(umap_cytof$data, col = "CD4", title = "CyTOF CD4")
plot_markers(umap_cytof$data, markers = c("CD14", "CD16", "CD68", "HLADR", "CD3", "CD4", "CD8a", "CD19", "CD20", "CD56", "CD38"), show_legend = F)
plot_markers(umap_cytof$data, markers = sort(markers), show_legend = F)


cyDefine::plot_heatmap(umap_cytof$data, population_col = "label", markers_to_plot = sort(markers))
```

| Cell Type | Markers |
|-----------|---------|
| CD4 T cells | CD3+, CD4+ |
| CD8 T cells | CD3+, CD8a+ |
| B cells | CD19+, CD20+, CD79b+, HLADR+ |
| NK cells | CD56+ |
| NKT cells | CD56+, CD3+ |
| CD16- monocytes | CD14+, CD16- |
| CD16+ monocytes | CD16+ |
| DCs | HLADR+, CD68+ |
| DN T cells | CD3+, CD4-, CD8a- |
| DP T cells | CD3+, CD4+, CD8a+ |

```{r}
annotation <- c(
  "1" = "CD8 T cell", 
  "2" = "CD8 T cell", 
  "3" = "CD8 T cell", 
  "4" = "CD8 T cell", 
  "5" = "DP T cell", 
  "6" = "DN T cell", 
  "7" = "CD4 T cell", 
  "8" = "CD4 T cell",
  "9" = "NKT cell",
  "10" = "NK cell",
  "11" = "CD4 T cell", 
  "12" = "CD4 T cell", 
  "13" = "dendritic cell", 
  "14" = "unassigned", 
  "15" = "dendritic cell",
  "16" = "DN T cell",
  "17" = "CD16- monocyte",
  "18" = "B cell")

cytof <- cytof |> 
  mutate(label = stringr::str_remove_all(label, "cluster_"),
         batch = "CyTOF",
         celltype = annotation[label])

umap_cytof$data$celltype <- annotation[stringr::str_remove_all(umap_cytof$data$label, "cluster_")]

unique_cells <- c(as.character(seu$celltype), cytof$celltype) |> unique() |> sort()
celltype_colors <- cyDefine::get_distinct_colors(unique_cells)
celltype_colors["dendritic cell"] <- "#AA0A0A"

plot_cytof_exprs <- plot_embedding(umap_cytof$data, col = "celltype", title = "CyTOF", colors = celltype_colors, legend_title = "Cell type")
plot_cytof_exprs
```


### Convert protein to gene

I convert protein names to gene names to find the common features between the two datasets.

```{r convert_genename}

protein_gene <- data.frame(
  Protein = c("CD45", "CD11b", "CD79b", "CD11c", "IgD", "CD16", "CD25", "CD43", "CD103", "CD45RA",
              "Tbet", "FoxP3", "CD161", "CD8a", "CD49a", "CD3", "CD20", "CD45RO", "HLADR", "CD56"),
  Gene_Symbol = c("PTPRC", "ITGAM", "CD79B", "ITGAX", "IGHD", "FCGR3A", "IL2RA", "SPN", "ITGAE",
                 "PTPRC", "TBX21", "FOXP3", "KLRB1", "CD8A", "ITGA1", "CD3E", "MS4A1", "PTPRC",
                 "HLA-DRA", "NCAM1")
)


if ("CD45RO" %in% colnames(cytof)) {
  cd45 <- pmax(cytof$CD45, cytof$CD45RA, cytof$CD45RO)
  cytof <- cytof |> 
    select(-starts_with("CD45")) |> 
    mutate(CD45 = cd45)
  markers <- get_markers(cytof)
}

genes_to_rename <- rownames(seu)[rownames(seu) %in% protein_gene$Gene_Symbol]
rownames(seu)[rownames(seu) %in% genes_to_rename] <- protein_gene$Protein[match(genes_to_rename, protein_gene$Gene_Symbol)]

```

## Run scennep

I run single-cell nearest-neighbor pseudobulking (scennep) using cosine distance.
The umap demonstrates the impact of running scennep on the data.


```{r scennep}
## Run scennep
library(scennep)
rna <- scennep(
  seu,
  markers = markers, 
  mc.cores = 4,
  distance = "cosine",
  return_S4 = FALSE
)
```

### Merge datasets

```{r merge}
seed <- 447
rna <- as.data.frame(t(rna))
rna$celltype <- seu$celltype
rna$batch <- "RNA"
rna$sample <- "RNA"
cytof$batch <- "CyTOF"


# Merge datasets
set.seed(seed)
uncorrected <- rna |> 
  dplyr::select(dplyr::any_of(c(markers, non_markers))) |> 
  dplyr::bind_rows(cytof |> 
                     dplyr::select(dplyr::any_of(c(markers, non_markers)))) |> 
  dplyr::mutate(id = dplyr::row_number())
```

## Uncorrected

```{r uncorrected}
set.seed(seed)
uncor_batch <- uncorrected |> 
  dplyr::group_by(sample) |> 
  dplyr::slice_sample(n = nrow(rna)) |> 
  plot_umap(
  down_sample = F,
  title = "Uncorrected",
  legend_title = "Technology",
  metric = "cosine",
  markers = markers,
  col = "batch")

uncor_batch
```


## Batch correct
```{r correct}

system.time({
  suppressMessages({ # Hide per-cluster correction info
    corrected <- cyCombine(
      uncorrected,
      markers = markers,
      norm_method = "scale",
      cluster_method = "flowsom",
      distf = "cosine",
      # ref.batch = "CyTOF",
      seed = seed,
      xdim = c(1, 5),
      ydim = c(1, 5),
      mc.cores = 4,
      method = "ComBat"
    )
  })
})
```


## Plot results - Batch

```{r plot_batch}
set.seed(seed)
plot_batch <- corrected |> 
  dplyr::group_by(sample) |> 
  dplyr::slice_sample(n = 10000) |> 
  plot_umap(
  down_sample = F,
  markers = markers,
  title = "CyTOF + scRNA-seq", 
  return_data = T,
  metric = "cosine", 
  legend_title = "Technology",
  col = "batch")
# ggplot2::ggsave(plot_batch$plot, filename = "figs/05_cytof_rnaseq_batch.png")
# saveRDS(plot_batch$plot, "results/05_cytof_rnaseq_batch.rds")
plot_batch$plot
```

## Plot results - Cells

Here, I reuse the umap generated above to plot celltype labels on the two modalities side by side.

```{r plot_cells}
xlim <- c(min(plot_batch$data$UMAP1), max(plot_batch$data$UMAP1))
ylim <- c(min(plot_batch$data$UMAP2), max(plot_batch$data$UMAP2))

plot_cell <- 
  plot_embedding(
    dplyr::filter(plot_batch$data, batch == "CyTOF"),
    colors = celltype_colors, 
    xlim = xlim,
    ylim = ylim,
    title = "Integrated CyTOF",
    col = "celltype",
    legend_title = "Cell type") + 
  plot_embedding(
    dplyr::filter(plot_batch$data, batch == "RNA"),
    colors = celltype_colors, 
    xlim = xlim,
    ylim = ylim,
    title = "Integrated scRNA-seq scennep",
    col = "celltype",
    legend_title = "Cell type") + 
  patchwork::plot_layout(
    axes = "collect",
    axis_titles = "collect")

plot_cell
```

## Phenotype transfer with cyDefine

```{r cyDefine}
markers <- cyCombine::get_markers(corrected)
classified <- cyDefine::cyDefine(
  corrected |> dplyr::filter(batch == "CyTOF"),
  corrected |> dplyr::filter(batch == "RNA"),
  markers = markers,
  using_pbmc = FALSE,
  adapt_reference = FALSE,
  batch_correct = FALSE,
  seed = seed,
  identify_unassigned = TRUE,
  train_on_unassigned = FALSE,
  mtry = 10,
  num.threads = 4,
  num.trees = 1000
)

```

### Figure 4A - UMAP
```{r}
colors <- cyDefine::get_distinct_colors(sort(unique(classified$reference$celltype)), add_unassigned = TRUE)
p_umap <- plot_umap(
    classified$reference,
    classified$query,
    markers = markers,
    colors = colors, 
    ref_col = "celltype",
    query_col = "predicted_celltype",#"model_prediction",
    build_umap_on = "reference",
    title = c("Reference (CyTOF)", "Query (scRNA-seq)"),
    seed = seed)
# saveRDS(p_umap, "results/05_cytof_rnaseq_umap.rds")
# ggplot2::ggsave(p_umap, filename = "figs/05_cytof_rnaseq_umap.png")
p_umap 
```


### Figure 4B - Celltype abundances

```{r abundance}
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
p_abundance <- plot_abundance_comparison(
  classified, 
  ref_name = "CyTOF", 
  query_name = "scRNA-seq", 
  query_col = "predicted_celltype",
  CLR = TRUE)

# saveRDS(p_abundance, "results/05_cytof_rnaseq_abundance.rds")
# ggplot2::ggsave(p_abundance, filename = "figs/05_cytof_rnaseq_abundance.png")
p_abundance 
```

### Figure 4C - Marker correlation

```{r marker_correlation}
p_correlation <- plot_expression_correlation(classified, markers = markers, ref_name = "CyTOF", query_name = "scRNA-seq")

uncorrected$celltype <- c(as.character(classified$query$predicted_celltype), classified$reference$celltype)
p_correlation_raw <- plot_expression_correlation(uncorrected |> dplyr::filter(batch == "CyTOF"),
  uncorrected |> dplyr::filter(batch == "RNA"), query_col = "celltype", markers = markers, ref_name = "CyTOF", query_name = "scRNA-seq")

# saveRDS(p_correlation, "results/05_cytof_rnaseq_correlation.rds")
# ggplot2::ggsave(p_correlation, filename = "figs/05_cytof_rnaseq_correlation.png", height = 20, width = 20, units = "cm")
p_correlation


p_collected <- (( p_umap / p_abundance + theme(legend.position = "right")) | p_correlation ) +
  plot_layout(tag_level = 'new') + 
  plot_annotation(tag_levels = list(c("A", " ","B","C")))

p_collected



ggplot2::ggsave(p_collected, filename = "figs/05_cytof_rnaseq_abundance_correlation.png", height = 20, width = 35, units = "cm")
# ggplot2::ggsave(p_collected, filename = "figs/05_cytof_rnaseq_abundance_correlation.pdf", height = 20, width = 35, units = "cm")
```
