---
title: "CEACAM5"

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_format = "html_document",
      output_dir = "../results/html_docs",
    )
  })
---
<style>
  .scrollingplot{
      overflow-x:scroll;
      white-space: nowrap;
  }

  .scrollingplot img{
     max-width: 40cm;
  }

</style>

```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
setwd("../../")
source("cartcontent/scr/01_functions.R")
```

<h4>Description</h4>
Carcinoembryonic antigen-related cell adhesion molecule 5 ([CEACAM5](https://www.uniprot.org/uniprot/P06731)). Cell surface glycoprotein that plays a role in cell adhesion, intracellular signaling and tumor progression. Plays a role as an oncogene by promoting tumor progression; induces resistance to anoikis of colorectal carcinoma cells.   

<br> 

<h4>Clinical trials</h4> 
There are no clinical trials involving CEACAM5, however the following two clinical trials involve CEA (keep or change?). <br>
[NCT02349724](https://clinicaltrials.gov/ct2/show/NCT02349724) - Lung Cancer, Colorectal Cancer, Breast Cancer, Pancreatic Cancer <br>
[NCT03267173](https://clinicaltrials.gov/ct2/show/NCT03267173) - Pancreatic Cancer 

<br>

<h4>Target gene expression</h4>
The following plot describes the expression, in different tissues and TCGA cancers, of the different isoforms of the gene CEACAM5.

<div class="scrollingplot">
```{r , echo = FALSE, warning = FALSE, message = FALSE, fig.width = 80, fig.height = 40}
setwd("../../")
boxplot_isoforms_all_tissues(hgnc = "CEACAM5") # it creates the boxplot in the html and saves it as .pdf in the plot/isoform_expression folder
```
</div>
<br>

<h4>Target essentiality</h4>
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 25, fig.height= 15}
setwd("../../")
target_essentiality_fun(gene_name_hgnc = "CEACAM5") # it creates the boxplot in the html and saves it as .pdf in the plot/target_essentiality folder
```

<br>

<h4>Protein properties</h4>
The following plot describes the target topology, highlighting the position of the endo-, ecto- and transmembrane domains, as well as the position of a signal peptide, when present.
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 10, fig.height= 3}
setwd("../../")
plot_function("CEACAM5") # it creates the plot in the html and saves it as .pdf in the plot/protein_properties folder
```