---
title: "MSLN"

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_format = "html_document",
      output_dir = "../results/html_docs",
    )
  })
---
<style>
img {
    max-width: 30cm;
}
</style>
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
setwd("../../")
source("cartcontent/scr/00_functions.R")
```

<h4>Description</h4>
Mesothelin ([MSLN](https://www.uniprot.org/uniprot/Q13421)). It plays a role in cell adhesion. 

<br> 

<h4>Clinical trials</h4> 
[NCT02792114](https://clinicaltrials.gov/ct2/show/NCT02792114) - Breast Cancer <br>
[NCT03356795](https://clinicaltrials.gov/ct2/show/NCT03356795) - Cervical Cancer <br>
[NCT04503980](https://clinicaltrials.gov/ct2/show/NCT04503980) – Colorectal Cancer, Ovarian Cancer <br>
[NCT03638206](https://clinicaltrials.gov/ct2/show/NCT03638206) – Colorectal Cancer, Esophagus Cancer, Renal Carcinoma, Hepatoma, Lung Cancer, Mesothelioma, Ovarian Cancer, Pancreatic Cancer, Melanoma <br>
[NCT03941626](https://clinicaltrials.gov/ct2/show/NCT03941626) – Esophagus Cancer, Hepatoma <br>
[NCT02959151](https://clinicaltrials.gov/ct2/show/NCT02959151) – Hepatoma <br>
[NCT03054298](https://clinicaltrials.gov/ct2/show/NCT03054298) - Lung Adenocarcinoma, Ovarian Cancer <br>
[NCT04489862](https://clinicaltrials.gov/ct2/show/NCT04489862) - Mesothelioma <br>
[NCT02580747](https://clinicaltrials.gov/ct2/show/NCT02580747) – Mesothelioma, Ovarian Cancer, Pancreatic Cancer, Endometrial Cancer <br>
[NCT03814447](https://clinicaltrials.gov/ct2/show/NCT03814447) – Ovarian Cancer <br>
[NCT03497819](https://clinicaltrials.gov/ct2/show/NCT03497819) – Pancreatic Cancer <br>
[NCT03323944](https://clinicaltrials.gov/ct2/show/NCT03323944) - Pancreatic Cancer <br>
[NCT03638193](https://clinicaltrials.gov/ct2/show/NCT03638193) - Pancreatic Cancer <br>
[NCT02706782](https://clinicaltrials.gov/ct2/show/NCT02706782) - Pancreatic Cancer <br>
[NCT02465983](https://clinicaltrials.gov/ct2/show/NCT02465983) - Pancreatic Cancer <br>
[NCT03267173](https://clinicaltrials.gov/ct2/show/NCT03267173) - Pancreatic Cancer <br>

<br>

<h4>Target gene expression</h4>
The following plot describes the expression, in different tissues and TCGA cancers, of the different isoforms of the gene MSLN.
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 32, fig.height= 20}
setwd("../../")
boxplot_isoforms_all_tissues(hgnc = "MSLN") # it creates the boxplot in the html and saves it as .pdf in the plot/isoform_expression folder
```

<br>

<h4>Target essentiality</h4>
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 25, fig.height= 12}
setwd("../../")
target_essentiality_fun(gene_name_hgnc = "MSLN") # it creates the boxplot in the html and saves it as .pdf in the plot/target_essentiality folder
```

<br>

<h4>Protein properties</h4>
The following plot describes the target topology, highlighting the position of the endo-, ecto- and transmembrane domains, as well as the position of a signal peptide, when present. 
The last three rows describe the position of potential epitopes, as well as phosphorylation and glycosylation sites. 
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 10, fig.height= 4}
setwd("../../")
plot_function("MSLN") # it creates the plot in the html and saves it as .pdf in the plot/protein_properties folder
```