---
title: "TNFRSF17"

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_format = "html_document",
      output_dir = "../results/html_docs",
    )
  })
---
<style>
img {
    max-width: 30cm;
}
</style>
```{r, eval = FALSE, echo = FALSE, warning = FALSE, message = FALSE}
setwd("../../")
source("cartcontent/scr/00_functions.R")
```


<h4>Description</h4>
Tumor necrosis factor receptor superfamily member 17 ([TNFRSF17](https://www.uniprot.org/uniprot/Q02223)). Receptor for TNFSF13B/BLyS/BAFF and TNFSF13/APRIL. Promotes B-cell survival and plays a role in the regulation of humoral immunity. Activates NF-kappa-B and JNK.

<br> 

<h4>Clinical trials</h4>
[NCT03302403](https://clinicaltrials.gov/ct2/show/NCT03302403) - Hepatocellular Carcinoma, Pancreatic Cancer

<br>

<h4>Target gene expression</h4>
```{r, eval = F, echo = FALSE, warning = FALSE, message = F, fig.width = 32, fig.height= 20}
setwd("../../")
boxplot_isoforms_all_tissues(hgnc = "TNFRSF17") # it creates the boxplot in the html and saves it as .pdf in the plot/isoform_expression folder

# in the file expr_sub.Rdata, TNFRSF17 does not exist. There is instead ENSG00000172322 (CLEC12A) that is not among the TCGA CarT target. 

```

<br>

<h4>Target essentiality</h4>

<br>

<h4>Protein properties</h4>
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 10, fig.height= 4}
setwd("../../")
plot_function("TNFRSF17") # it creates the plot in the html and saves it as .pdf in the plot/protein_properties folder
```