---
title: "TNFRSF10B"

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_format = "html_document",
      output_dir = "../results/html_docs",
    )
  })
---
<style>
img {
    max-width: 30cm;
}
</style>
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
setwd("../../")
source("cartcontent/scr/00_functions.R")
```


<h4>Description</h4>
Tumor necrosis factor receptor superfamily member 10B ([TNFRSF10B](https://www.uniprot.org/uniprot/O14763)). Receptor for the cytotoxic ligand TNFSF10/TRAIL. Promotes the activation of NF-kappa-B. Essential for ER stress-induced apoptosis.

<br> 

<h4>Clinical trials</h4>
[NCT03638206](https://clinicaltrials.gov/ct2/show/NCT03638206) - Colorectal Cancer, Esophagus Cancer, Renal Carcinoma, Hepatoma, Lung Cancer, Mesothelioma, Ovarian Cancer, Pancreatic Cancer, Melanoma

<br>

<h4>Target gene expression</h4>
The following plot describes the expression, in different tissues and TCGA cancers, of the different isoforms of the gene TNFRSF10B.
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 32, fig.height= 20}
setwd("../../")
boxplot_isoforms_all_tissues(hgnc = "TNFRSF10B") # it creates the boxplot in the html and saves it as .pdf in the plot/isoform_expression folder
```
<
<br>

<h4>Target essentiality</h4>
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 25, fig.height= 12}
setwd("../../")
target_essentiality_fun(gene_name_hgnc = "TNFRSF10B") # it creates the boxplot in the html and saves it as .pdf in the plot/target_essentiality folder
```

<br>

<h4>Protein properties</h4>
The following plot describes the target topology, highlighting the position of the endo-, ecto- and transmembrane domains, as well as the position of a signal peptide, when present. 
The last three rows describe the position of potential epitopes, as well as phosphorylation and glycosylation sites. 
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 10, fig.height= 4}
setwd("../../")
plot_function("TNFRSF10B") # it creates the plot in the html and saves it as .pdf in the plot/protein_properties folder
```