---
title: "CAR-T Target Assessment Workflow - coding"
author: "Giorgia Moranzoni"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      eval = TRUE, fig.width = 12, fig.height = 8)
```

# Introduction
Chimeric Antigen Receptor T-cell (CAR-T) therapy represents a revolutionary approach to cancer treatment, where a patient's T-cells are genetically modified to recognize and attack cancer cells. The success of CAR-T therapy critically depends on identifying suitable target antigens—proteins expressed on cancer cells that can serve as targets for the engineered T-cells.

This vignette exemplifies a comprehensive computational workflow for CAR-T target assessment at the isoform level. We demonstrate this workflow using **ERBB2** (HER2/neu) as a case study, systematically evaluating isoforms based on subcellular localization, membrane topology, epitope retention, and expression patterns. **ERBB"** is a receptor tyrosine kinase overexpressed in approximately 25-30% of breast cancers [Slamon, D.J. et al.](https://pubmed.ncbi.nlm.nih.gov/2470152/) and an established target for monoclonal antibody therapy (trastuzumab) [Yoon, J. et al.](https://www.nature.com/articles/s41571-024-00924-9#citeas). 

## Required Packages

```{r load_packages, eval=FALSE}
library(biomaRt)
library(Biostrings)
library(ggnewscale)
library(patchwork)
library(tidyverse) 

options(timeout = 300)
theme_set(theme_light())

```

# Phase 1: Isoform Discovery

## Step 1: Retrieve Isoform Sequences

Download protein sequences for all HER2 isoforms from Ensembl (GRCh38, release 115). We retrieve sequences for all protein-coding transcripts to ensure comprehensive coverage.

```{r retrieve_sequences, eval=FALSE}
# Connect to Ensembl
ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")

# Get protein-coding ERBB2 transcripts
get_gene_transcripts <- function(gene_symbol, mart) {
  getBM(
    attributes = c("ensembl_gene_id", "ensembl_transcript_id", 
                   "transcript_biotype", "transcript_length"),
    filters = "hgnc_symbol",
    values = gene_symbol,
    mart = mart
  ) %>%
    filter(transcript_biotype == "protein_coding") %>%
    arrange(desc(transcript_length))
}

her2_transcripts <- get_gene_transcripts("ERBB2", ensembl)

# Retrieve protein sequences
get_protein_sequences <- function(transcript_ids, mart) {
  seqs <- getBM(
    attributes = c("ensembl_transcript_id", "peptide"),
    filters = "ensembl_transcript_id",
    values = transcript_ids,
    mart = mart
  )
  
  sequences <- AAStringSet(seqs$peptide)
  names(sequences) <- seqs$ensembl_transcript_id
  sequences[width(sequences) > 0]  # Remove empty sequences
}

her2_sequences_all <- get_protein_sequences(her2_transcripts$ensembl_transcript_id, ensembl)
print(her2_sequences_all)
```

## Step 2: Download Expression Data

We use pre-processed TCGA/GTEx expression data from Xena Browser. This dataset provides transcript-level expression across cancer and normal tissues. Note: Ensembl's September 2025 update added ~120,000 new transcripts not yet reflected in available expression databases, so some isoforms are absent from the expression data.

```{r download_expression, eval=FALSE}
# File paths
expr_file <- "data/TcgaTargetGtex_rsem_isoform_tpm.gz"
filtered_file <- "results/her2_isoforms_filtered.txt"
output_rds <- "results/her2_expr_filtered.rds"

# HER2 transcripts to extract
her2_transcripts <- names(her2_sequences_all)

# Filter expression file line-by-line (efficient for large files)
con <- gzfile(expr_file, "r")
header <- readLines(con, n = 1)
writeLines(header, filtered_file)

line_count <- 0
matched_count <- 0

while (TRUE) {
  lines <- readLines(con, n = 1000)
  if (length(lines) == 0) break
  
  line_count <- line_count + length(lines)
  
  # Extract HER2 transcript lines
  pattern <- paste0("^(", paste(her2_transcripts, collapse = "|"), ")(\\.|\\t)")
  matching_lines <- grep(pattern, lines, value = TRUE)
  
  if (length(matching_lines) > 0) {
    write(matching_lines, file = filtered_file, append = TRUE)
    matched_count <- matched_count + length(matching_lines)
  }
}

close(con)

# Read filtered data and save as RDS
her2_expr <- read_tsv(filtered_file, show_col_types = FALSE)
saveRDS(her2_expr, output_rds)
```

## Step 3: Identify Available Transcripts

Identify which HER2 isoforms have expression data and filter sequences accordingly.

```{r identify_transcripts, eval=FALSE}
# Load pre-filtered expression data
her2_expr <- readRDS("../../results/HER2_analysis/her2_expr_filtered.rds")

# Extract transcript IDs (remove version numbers)
her2_available_transcripts <- her2_expr %>% 
  mutate(transcript_clean = str_remove(sample, "\\..*")) %>%
  pull(transcript_clean) %>%
  unique()

# Filter sequences to those with expression data
sequence_names_clean <- str_remove(names(her2_sequences_all), "\\..*")
keep_indices <- sequence_names_clean %in% her2_available_transcripts
her2_sequences_filtered <- her2_sequences_all[keep_indices]

cat("Sequences with expression data:", length(her2_sequences_filtered), "\n")
```

## Step 4: Export for External Tools

Save sequences in FASTA format for external prediction tools (DeepLoc2 for subcellular localization, DeepTMHMM for membrane topology).

```{r export_fasta, eval=FALSE}
fasta_file <- "../../data/her2_isoforms_filtered.fasta"
writeXStringSet(her2_sequences_filtered, filepath = fasta_file)

cat("✓ FASTA saved:", fasta_file, "\n\n")
cat("Upload to:\n")
cat("  • DeepLoc2: https://services.healthtech.dtu.dk/services/DeepLoc-2.1/\n")
cat("  • DeepTMHMM: https://dtu.biolib.com/DeepTMHMM\n")
```

# Phase 2: Structural Characterization

## Step 5: Align Isoforms with MAFFT

Align sequences using MAFFT with the `--localpair` algorithm, which is optimized for sequences with large insertions/deletions from alternative splicing. This ensures proper alignment of conserved functional domains.

```{r align_sequences, eval=FALSE}
temp_input <- tempfile(fileext = ".fasta")
temp_output <- tempfile(fileext = ".fasta")

writeXStringSet(her2_sequences_filtered, temp_input)

# MAFFT alignment optimized for isoforms
system2("mafft", 
        args = c("--localpair", "--maxiterate", "1000", "--quiet", temp_input),
        stdout = temp_output)

her2_aligned <- readAAStringSet(temp_output)
unlink(c(temp_input, temp_output))

cat("Alignment width:", unique(width(her2_aligned)), "positions\n")
# writeXStringSet(her2_aligned, "../../results/HER2_analysis/her2_aligned_mafft.fasta")
```

## Step 6: Import and Process Predictions

Import external prediction results and perform integrated analysis: classify isoforms by subcellular localization, map membrane topology to aligned sequences, and identify therapeutic epitope retention across all isoforms.

```{r process_predictions, eval=FALSE}
# ===== Import DeepLoc2 Results =====
deeploc_results <- read_csv("../../cartcontent/results/HER2_analysis/HER2_deeploc21_filtered.csv", 
                             show_col_types = FALSE)
deeploc_filtered <- deeploc_results %>%
  mutate(transcript_clean = str_remove(Protein_ID, "\\..*")) %>%
  filter(transcript_clean %in% her2_available_transcripts)

# ===== Import DeepTMHMM Results =====
parse_deeptmhmm <- function(file_path) {
  lines <- readLines(file_path)
  results <- list()
  i <- 1
  
  while (i <= length(lines)) {
    if (startsWith(lines[i], ">")) {
      transcript_id <- str_extract(lines[i], "(?<=>)[^ ]+")
      sequence <- lines[i + 1]
      topology <- lines[i + 2]
      
      results[[transcript_id]] <- list(
        transcript_id = transcript_id,
        sequence = sequence,
        topology = topology
      )
      i <- i + 3
    } else {
      i <- i + 1
    }
  }
  return(results)
}

deeptmhmm_results <- parse_deeptmhmm(
  "../../cartcontent/results/HER2_analysis/HER2_deeptmhmm_predicted_topologies_filtered.3line"
)

# ===== Classify Isoforms by Localization =====
transcript_classification <- deeploc_filtered %>%
  dplyr::select(Protein_ID, `Cell membrane`, Extracellular) %>%
  mutate(
    transcript_clean = str_remove(Protein_ID, "\\..*"),
    classification = case_when(
      `Cell membrane` > Extracellular ~ "Membrane-bound",
      Extracellular > `Cell membrane` ~ "Secreted",
      TRUE ~ "Ambiguous"
    )
  )

classification_colors <- c(
  "Membrane-bound" = "#E74C3C",
  "Secreted" = "#3498DB",
  "Ambiguous" = "#95A5A6"
)

# ===== Map Topology to Alignment =====
map_topology_to_alignment <- function(deeptmhmm_results, aligned_sequences) {
  map_dfr(names(deeptmhmm_results), function(tid) {
    topology <- deeptmhmm_results[[tid]]$topology
    aligned_seq <- as.character(aligned_sequences[tid])
    
    topology_chars <- strsplit(topology, "")[[1]]
    aligned_chars <- strsplit(aligned_seq, "")[[1]]
    aligned_topology <- character(length(aligned_chars))
    
    # Transfer topology to aligned positions
    original_pos <- 1
    for (aligned_pos in 1:length(aligned_chars)) {
      if (aligned_chars[aligned_pos] == "-") {
        aligned_topology[aligned_pos] <- "-"
      } else {
        if (original_pos <= length(topology_chars)) {
          aligned_topology[aligned_pos] <- topology_chars[original_pos]
        }
        original_pos <- original_pos + 1
      }
    }
    
    data.frame(
      position = 1:length(aligned_chars),
      aa = aligned_chars,
      topology = aligned_topology,
      transcript_id = tid
    ) %>%
      mutate(
        topology_type = case_when(
          topology == "S" ~ "Signal peptide",
          topology == "O" ~ "Extracellular",
          topology == "M" ~ "Transmembrane",
          topology == "I" ~ "Intracellular",
          topology == "-" ~ "Alignment gap",
          TRUE ~ "Unknown"
        )
      )
  })
}

aligned_segments <- map_topology_to_alignment(deeptmhmm_results, her2_aligned)

# ===== Map Epitopes Across Isoforms =====
# Therapeutic antibody epitope positions in canonical HER2
epitope_positions <- list(
  P = list(c(257, 267), c(308, 318)),  # Pertuzumab
  T = list(c(580, 582), c(593, 595), c(610, 625))  # Trastuzumab
)

map_epitopes_from_alignment <- function(aligned_seqs, canonical_name, epitope_pos) {
  canonical_seq <- as.character(aligned_seqs[[canonical_name]])
  epitope_map_list <- list()
  
  for(seq_name in names(aligned_seqs)) {
    isoform_seq <- as.character(aligned_seqs[[seq_name]])
    canonical_ungapped_pos <- 0
    epitope_positions_aligned <- list()
    
    for(i in 1:nchar(canonical_seq)) {
      canon_char <- substr(canonical_seq, i, i)
      iso_char <- substr(isoform_seq, i, i)
      
      if(canon_char != "-") canonical_ungapped_pos <- canonical_ungapped_pos + 1
      
      # Check if position is in epitope and isoform has residue
      for(epitope_name in names(epitope_pos)) {
        for(region in epitope_pos[[epitope_name]]) {
          if(canonical_ungapped_pos >= region[1] && 
             canonical_ungapped_pos <= region[2] &&
             iso_char != "-") {
            epitope_positions_aligned[[length(epitope_positions_aligned) + 1]] <- data.frame(
              epitope = epitope_name,
              position = i,
              canonical_pos = canonical_ungapped_pos
            )
          }
        }
      }
    }
    
    if(length(epitope_positions_aligned) > 0) {
      epitope_map_list[[seq_name]] <- bind_rows(epitope_positions_aligned)
    }
  }
  
  bind_rows(epitope_map_list, .id = "transcript_id")
}

epitope_mapping <- map_epitopes_from_alignment(
  aligned_seqs = her2_aligned,
  canonical_name = "ENST00000269571",
  epitope_pos = epitope_positions
)

# Summarize epitope retention
epitope_summary <- epitope_mapping %>%
  group_by(transcript_id, epitope) %>%
  summarise(n_positions = n(), .groups = "drop") %>%
  pivot_wider(names_from = epitope, values_from = n_positions, values_fill = 0)

print(epitope_summary)
```

## Step 7: Visualize Integrated Results

Create comprehensive visualization combining subcellular localization (DeepLoc2), membrane topology (DeepTMHMM), and epitope locations in a single integrated figure.

```{r visualize_integrated, eval=FALSE, fig.width=20, fig.height=8}
# Define isoform display order
transcript_order <- c(
  "ENST00000580074",
  "ENST00000578709",
  "ENST00000584601",
  "ENST00000584450",
  "ENST00000584014",
  "ENST00000582818",
  "ENST00000578502",  # Longest
  "ENST00000578199",  # Secreted
  "ENST00000445658",  # Truncated
  "ENST00000269571"   # Canonical
)

# ===== DeepLoc2 Heatmap =====
localization_cols <- c("Cell membrane", "Cytoplasm", "Endoplasmic reticulum",
                       "Extracellular", "Golgi apparatus", "Lysosome/Vacuole",
                       "Mitochondrion", "Nucleus", "Peroxisome", "Plastid")

plot_data_combined <- deeploc_filtered %>%
  mutate(Protein_ID = factor(Protein_ID, levels = transcript_order)) %>%
  dplyr::select(Protein_ID, all_of(localization_cols)) %>%
  pivot_longer(cols = -Protein_ID, names_to = "localization", values_to = "probability")

deeploc_heatmap <- ggplot(plot_data_combined, 
                          aes(x = localization, y = Protein_ID, fill = probability)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradientn(
    colors = c("#4A90B8", "#5E9DC3", "#FFFACD", "#F26D5E", "#D73027"),
    values = c(0, 0.2, 0.5, 0.8, 1),
    limits = c(0, 1)
  ) +
  labs(title = "HER2 Isoforms", x = NULL, y = NULL, fill = "Probability") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
    axis.text.y = element_text(size = 16, family = "mono"),
    plot.title = element_text(size = 18, face = "bold"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 16),
    legend.position = "top",
    panel.grid = element_blank()
  )

# ===== Topology Plot with Epitopes =====
plot_topology_with_epitopes <- function(aligned_segments, epitope_data, transcript_order) {
  topology_colors <- c(
    "Signal peptide" = "#C41C24",
    "Extracellular" = "#FFB20F",
    "Transmembrane" = "#18848C",
    "Intracellular" = "#96BDC6",
    "Alignment gap" = "#EDE7E3"
  )
  
  epitope_colors <- c("P" = "#55a630", "T" = "#ff5d8f")
  
  aligned_segments <- aligned_segments %>%
    mutate(transcript_id = factor(transcript_id, levels = transcript_order))
  
  # Base topology plot
  p <- ggplot(aligned_segments, aes(x = position, y = transcript_id, fill = topology_type)) +
    geom_tile(height = 0.8, width = 1) +
    scale_fill_manual(values = topology_colors, name = "Topology") +
    scale_x_continuous(expand = c(0, 0), breaks = seq(0, max(aligned_segments$position), 400)) +
    labs(x = "Amino acid position (aligned)", y = NULL) +
    theme_minimal() +
    theme(
      axis.text.y = element_blank(),
      axis.text.x = element_text(size = 16),
      axis.title.x = element_text(size = 18),
      legend.position = "right",
      legend.text = element_text(size = 16),
      legend.title = element_text(size = 18),
      panel.grid = element_blank()
    )
  
  # Add epitope overlay
  if(!is.null(epitope_data) && nrow(epitope_data) > 0) {
    epitope_data <- epitope_data %>%
      mutate(transcript_id = factor(transcript_id, levels = transcript_order))
    
    p <- p +
      new_scale_fill() +
      geom_tile(data = epitope_data,
                aes(x = position, y = transcript_id, fill = epitope),
                height = 0.8, width = 1, alpha = 0.9, inherit.aes = FALSE) +
      scale_fill_manual(
        values = epitope_colors,
        name = "Epitope",
        labels = c("P" = "Pertuzumab", "T" = "Trastuzumab")
      )
  }
  
  return(p)
}

topology_plot <- plot_topology_with_epitopes(aligned_segments, epitope_mapping, transcript_order)

# ===== Combined Plot =====
combined_plot <- deeploc_heatmap + topology_plot + 
  plot_layout(widths = c(1, 1.5))

print(combined_plot)

# ggsave("../../results/HER2_analysis/figures/her2_localization_topology_epitopes.pdf",
#        combined_plot, width = 20, height = 8)
```

# Phase 3: Expression Analysis

## Step 8: Prepare Expression Data

Filter to targetable isoforms (membrane-bound with retained epitopes) and merge with tissue annotations from TCGA/GTEx.

```{r prepare_expression, eval=FALSE}
# Targetable isoforms based on topology and epitope analysis
targetable_isoforms <- c("ENST00000269571", "ENST00000584601", "ENST00000584450", 
                         "ENST00000584014", "ENST00000578502", "ENST00000578199", 
                         "ENST00000445658", "ENST00000582818")

# Load tissue annotations
phenotype <- read_tsv("../../data/HER2_TCGA_GTEX_category_2026.txt", show_col_types = FALSE) %>%
  separate(TCGA_GTEX_main_category, 
           into = c("dataset", "tissue_or_disease"), 
           sep = " ", extra = "merge") %>%
  mutate(tissue_or_disease = tolower(tissue_or_disease))

# Transform and filter expression data
her2_expr_long <- her2_expr %>%
  pivot_longer(cols = -sample, names_to = "sample_id", values_to = "log2_tpm001") %>%
  dplyr::rename(transcript_id = sample) %>%
  mutate(
    transcript_clean = str_remove(transcript_id, "\\..*"),
    # Convert log2(TPM + 0.001) to log2(TPM + 1)
    tpm = 2^log2_tpm001 - 0.001,
    tpm = pmax(tpm, 0),
    log2_tpm_plus1 = log2(tpm + 1)
  ) %>%
  filter(transcript_clean %in% targetable_isoforms)

# Merge with phenotype and classification
her2_expr_final <- her2_expr_long %>%
  left_join(phenotype, by = c("sample_id" = "sample")) %>%
  left_join(transcript_classification, by = "transcript_clean") %>%
  filter(!is.na(dataset))
```

## Step 9: Visualize Expression Patterns

Compare expression across normal tissues (GTEx) and cancer types (TCGA) to assess tumor selectivity and on-target/off-tumor toxicity risk.

```{r visualize_expression, eval=FALSE, fig.width=16, fig.height=8}
# Categorize normal tissues by clinical importance
tissue_categories <- list(
  "Critical" = c("brain", "blood", "blood vessel", "heart", "liver", "lung", "nerve"),
  "Important" = c("pancreas", "kidney", "stomach", "colon", "small intestine", "bladder"),
  "Others" = c("spleen", "uterus", "ovary", "testis", "prostate", "breast", "skin")
)

tissue_category_df <- data.frame(
  tissue_or_disease = unlist(tissue_categories),
  category = rep(names(tissue_categories), sapply(tissue_categories, length))
) %>%
  mutate(category = factor(category, levels = c("Critical", "Important", "Others")))

# ===== GTEx Data =====
gtex_data <- her2_expr_final %>%
  filter(dataset == "GTEX") %>%
  left_join(tissue_category_df, by = "tissue_or_disease") %>%
  filter(!is.na(category)) %>%
  arrange(category, transcript_clean) %>%
  group_by(category) %>%
  mutate(position_within_cat = as.numeric(factor(transcript_clean))) %>%
  ungroup() %>%
  mutate(
    gap_offset = case_when(
      category == "Critical" ~ 0,
      category == "Important" ~ 9,
      category == "Others" ~ 18
    ),
    x_position = gap_offset + position_within_cat
  )
gtex_data <- gtex_data %>%
  mutate(
    group_label = paste0(category, "_", transcript_clean),
  )

# ===== TCGA Data =====
tcga_cancers <- c("breast invasive carcinoma", "stomach adenocarcinoma", "glioblastoma multiforme")

tcga_data <- her2_expr_final %>%
  filter(dataset == "TCGA", tissue_or_disease %in% tcga_cancers) %>%
  arrange(tissue_or_disease, transcript_clean) %>%
  group_by(tissue_or_disease) %>%
  mutate(position_within_disease = as.numeric(factor(transcript_clean))) %>%
  ungroup() %>%
  mutate(
    gap_offset = case_when(
      tissue_or_disease == "breast invasive carcinoma" ~ 0,
      tissue_or_disease == "stomach adenocarcinoma" ~ 9,
      tissue_or_disease == "glioblastoma multiforme" ~ 18
    ),
    x_position = gap_offset + position_within_disease + 27 # Add offset to place after GTEx (26 + 1 gap)
  )

tcga_data <- tcga_data %>%
  mutate(
    group_label = paste0(tissue_or_disease, "_", transcript_clean),
  )

# ===== Combined Visualization =====
# Combine datasets
combined_data <- bind_rows(
  gtex_data %>% 
    dplyr::select(transcript_clean, log2_tpm_plus1, classification, 
           dataset, group_label, x_position),
  tcga_data %>% 
    dplyr::select(transcript_clean, log2_tpm_plus1, classification, 
           dataset, group_label, x_position)
)

expression_plot <- ggplot(combined_data, 
                          aes(x = x_position, 
                              y = log2_tpm_plus1, 
                              fill = classification, 
                              group = interaction(dataset, group_label))) +
  geom_boxplot(outlier.shape = NA) +
  scale_fill_manual(values = classification_colors, name = "Localization") +
  scale_x_continuous(
    breaks = combined_data %>% 
      distinct(x_position, group_label) %>% 
      pull(x_position),
    labels = combined_data %>% 
      distinct(x_position, group_label) %>% 
      pull(group_label) 
  ) +
  coord_cartesian(ylim = c(0, 7.5)) +
  geom_vline(xintercept = 26.5, linetype = "dashed", color = "gray50", linewidth = 0.5) +
  # Add dataset labels at the top
  annotate("text", x = 13, y = max(combined_data$log2_tpm_plus1) * 1.05, 
           label = "GTEx (Normal Tissues)", fontface = "bold", size = 4) +
  annotate("text", x = 40, y = max(combined_data$log2_tpm_plus1) * 1.05, 
           label = "TCGA (Cancer Types)", fontface = "bold", size = 4) +
  labs(
    title = "HER2 Isoform Expression Across Normal Tissues and Cancer Types",
    subtitle = "GTEx tissues grouped by essentiality, followed by TCGA cancer types",
    x = "Tissue/Cancer Type_Isoform",
    y = "Expression (log2(TPM + 1))"
  ) +
  theme_minimal() +
  theme(
    plot.margin = margin(t = 20, r = 10, b = 10, l = 40),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    legend.position = "right",
    legend.text = element_text(size = 10)
  )

print(expression_plot)

# ggsave("../../results/HER2_analysis/figures/her2_expression_gtex_tcga.pdf",
#        expression_plot, width = 16, height = 8)
```

## Step 10: Off-target binding assessment

### Sequence-level analysis (BLOSUM62)

```{r, eval=FALSE}
ensembl_family <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl", 
                              mirror = "uswest")

genes <- c("ERBB2", "ERBB3", "ERBB4", "EGFR")

canonical_ids <- getBM(
  attributes = c("hgnc_symbol", "ensembl_transcript_id", "transcript_is_canonical"),
  filters = "hgnc_symbol",
  values = genes,
  mart = ensembl_family
) %>%
  filter(transcript_is_canonical == 1)

seqs <- getBM(
  attributes = c("ensembl_transcript_id", "peptide"),
  filters = "ensembl_transcript_id",
  values = canonical_ids$ensembl_transcript_id,
  mart = ensembl_family
) %>%
  filter(peptide != "") %>%
  left_join(canonical_ids %>% select(hgnc_symbol, ensembl_transcript_id),
            by = "ensembl_transcript_id")

family_sequences <- AAStringSet(seqs$peptide)
names(family_sequences) <- paste0(seqs$hgnc_symbol, "_", seqs$ensembl_transcript_id)

temp_input <- tempfile(fileext = ".fasta")
temp_output <- tempfile(fileext = ".fasta")
writeXStringSet(family_sequences, temp_input)

system2("mafft",
        args = c("--localpair", "--maxiterate", "1000", "--quiet", temp_input),
        stdout = temp_output)

family_aligned <- readAAStringSet(temp_output)
unlink(c(temp_input, temp_output))

# Epitope positions in canonical HER2
epitope_positions <- list(
  P = list(c(257, 267), c(308, 318)),               # Pertuzumab
  T = list(c(580, 582), c(593, 595), c(610, 625))   # Trastuzumab
)

her2_name <- names(family_aligned)[str_detect(names(family_aligned), "ERBB2")]

epitope_mapping <- map_epitopes_from_alignment(
  aligned_seqs = family_aligned,
  canonical_name = her2_name,
  epitope_pos = epitope_positions
)

her2_seq <- strsplit(as.character(family_aligned[[her2_name]]), "")[[1]]

substitution_table <- epitope_mapping %>%
  mutate(
    aa = map2_chr(transcript_id, position, function(tid, pos) {
      strsplit(as.character(family_aligned[[tid]]), "")[[1]][pos]
    }),
    her2_aa = her2_seq[position]
  ) %>%
  filter(transcript_id != her2_name) %>%
  select(transcript_id, epitope, canonical_pos, her2_aa, aa) %>%
  mutate(changed = aa != her2_aa)

data(BLOSUM62)

substitution_table_scored <- substitution_table %>%
  filter(aa != "-") %>%
  rowwise() %>%
  mutate(blosum62 = BLOSUM62[her2_aa, aa]) %>%
  ungroup()

# Gaps (missing residues) are scored separately, not by BLOSUM62
gap_table <- epitope_mapping %>%
  filter(transcript_id == her2_name) %>%
  select(epitope, canonical_pos, position) %>%
  crossing(transcript_id = setdiff(names(family_aligned), her2_name)) %>%
  mutate(
    aa = map2_chr(transcript_id, position, function(tid, pos) {
      strsplit(as.character(family_aligned[[tid]]), "")[[1]][pos]
    })
  ) %>%
  filter(aa == "-") %>%
  mutate(her2_aa = her2_seq[position], changed = TRUE, blosum62 = NA) %>%
  select(transcript_id, epitope, canonical_pos, her2_aa, aa, changed, blosum62)

full_substitution_table <- bind_rows(substitution_table_scored, gap_table) %>%
  arrange(transcript_id, epitope, canonical_pos)

full_summary <- full_substitution_table %>%
  filter(transcript_id != her2_name) %>%
  mutate(protein = case_when(
    str_detect(transcript_id, "ERBB2") ~ "HER2",
    str_detect(transcript_id, "ERBB3") ~ "HER3",
    str_detect(transcript_id, "ERBB4") ~ "HER4",
    str_detect(transcript_id, "EGFR")  ~ "EGFR"
  )) %>%
  group_by(protein, epitope) %>%
  summarise(
    substitutions = sum(changed & aa != "-"),
    n_negative_blosum = sum(blosum62 < 0, na.rm = TRUE),
    gaps = sum(aa == "-"),
    mean_blosum = round(mean(blosum62[changed & aa != "-"], na.rm = TRUE), 2),
    .groups = "drop"
  )
```

### Structural analysis (backbone RMSD)
```{bash, eval = FALSE}
# RMSD calculated externally in PyMOL (Cα atoms only, no outlier rejection,
# cycles=0). The script below is shown for reproducibility but is not run
# as part of this vignette — results are hardcoded in rmsd_table.

# PyMOL: Pertuzumab epitope RMSD
fetch 1S78
fetch 1M6B
fetch 2AHX
fetch 1NQL

remove solvent
remove organic
remove 1M6B and chain B
remove 2AHX and chain B
remove 1NQL and chain B

hide everything
show cartoon, all

select ep_her2, 1S78 and chain A and (resi 235-245 or resi 286-296)
select ep_her3, 1M6B and chain A and (resi 229-239 or resi 279-289)
select ep_her4, 2AHX and chain A and (resi 226-236 or resi 276-286)
select ep_egfr, 1NQL and chain A and (resi 229-239 or resi 280-290)

align 1M6B and chain A and (resi 229-239 or resi 279-289) and name CA, 1S78 and chain A and (resi 235-245 or resi 286-296) and name CA, cycles=0
align 2AHX and chain A and (resi 226-236 or resi 276-286) and name CA, 1S78 and chain A and (resi 235-245 or resi 286-296) and name CA, cycles=0
align 1NQL and chain A and (resi 229-239 or resi 280-290) and name CA, 1S78 and chain A and (resi 235-245 or resi 286-296) and name CA, cycles=0


# PyMOL: Trastuzumab epitope RMSD
fetch 1N8Z
fetch 1M6B
fetch 2AHX
fetch 1NQL

remove solvent
remove organic
remove 1M6B and chain B
remove 2AHX and chain B
remove 1NQL and chain B

hide everything
show cartoon, all

select epitope_her2, 1N8Z and chain C and (resi 558-560 or resi 571-573 or resi 588-603)
select epitope_her3, 1M6B and chain A and (resi 550-552 or resi 563-565 or resi 580-593)
select epitope_her4, 2AHX and chain A and (resi 548-550 or resi 561-563 or resi 578-591)
select epitope_egfr, 1NQL and chain A and (resi 551-553 or resi 564-566 or resi 581-595)

align 1M6B and chain A and (resi 550-552 or resi 563-565 or resi 580-593) and name CA, 1N8Z and chain C and (resi 558-560 or resi 571-573 or resi 588-603) and name CA, cycles=0
align 2AHX and chain A and (resi 548-550 or resi 561-563 or resi 578-591) and name CA, 1N8Z and chain C and (resi 558-560 or resi 571-573 or resi 588-603) and name CA, cycles=0
align 1NQL and chain A and (resi 551-553 or resi 564-566 or resi 581-595) and name CA, 1N8Z and chain C and (resi 558-560 or resi 571-573 or resi 588-603) and name CA, cycles=0
```


```{r, eval=FALSE}
rmsd_table <- tibble(
  epitope = c(rep("P", 3), rep("T", 3)),
  protein = rep(c("HER3", "HER4", "EGFR"), 2),
  backbone_rmsd = c(0.871, 0.800, 1.336, 0.587, 0.709, 0.620),
  rmsd_atoms = c(20, 20, 18, 5, 17, 15)
)
rmsd_table
```

### B-factor analysis
```{r, eval=FALSE}
base_path <- "../../cartcontent/results/HER2_analysis/Pymol"

parse_bfactors <- function(pdb_file, resi_ranges, protein_name) {
  lines <- readLines(pdb_file)
  atom_lines <- lines[startsWith(lines, "ATOM")]
  
  tibble(
    resn = str_trim(substr(atom_lines, 18, 20)),
    resi = as.integer(str_trim(substr(atom_lines, 23, 26))),
    atom = str_trim(substr(atom_lines, 13, 16)),
    bfactor = as.numeric(substr(atom_lines, 61, 66))
  ) %>%
    filter(atom == "CA", resi %in% resi_ranges) %>%
    mutate(protein = protein_name) %>%
    select(protein, resi, resn, bfactor)
}

# Pertuzumab epitope B-factors (HER2 excluded — antibody-bound structure)
pert_bfactors <- bind_rows(
  parse_bfactors(file.path(base_path, "1S78_her2.pdb"), c(235:245, 286:296), "HER2"),
  parse_bfactors(file.path(base_path, "1M6B_her3.pdb"), c(229:239, 279:289), "HER3"),
  parse_bfactors(file.path(base_path, "2AHX_her4.pdb"), c(226:236, 276:286), "HER4"),
  parse_bfactors(file.path(base_path, "1NQL_egfr.pdb"), c(229:239, 280:290), "EGFR")
)

pert_bfactor_summary <- pert_bfactors %>%
  filter(protein != "HER2") %>%
  group_by(protein) %>%
  summarise(mean_bfactor = round(mean(bfactor), 2), .groups = "drop") %>%
  mutate(epitope = "P")

# Trastuzumab epitope B-factors (B = 0.0 residues excluded — unresolved density)
tras_bfactors <- bind_rows(
  parse_bfactors(file.path(base_path, "1N8Z_her2.pdb"), c(558:560, 571:573, 588:603), "HER2"),
  parse_bfactors(file.path(base_path, "1M6B_her3.pdb"), c(550:552, 563:565, 580:593), "HER3"),
  parse_bfactors(file.path(base_path, "2AHX_her4.pdb"), c(548:550, 561:563, 578:591), "HER4"),
  parse_bfactors(file.path(base_path, "1NQL_egfr.pdb"), c(551:553, 564:566, 581:595), "EGFR")
) %>%
  filter(bfactor > 0)

tras_expected <- c(HER2 = 22, HER3 = 20, HER4 = 20, EGFR = 21)

tras_bfactor_summary <- tras_bfactors %>%
  filter(protein != "HER2") %>%
  group_by(protein) %>%
  summarise(mean_bfactor = round(mean(bfactor), 2), n_resolved = n(), .groups = "drop") %>%
  mutate(
    n_expected = tras_expected[protein],
    n_unresolved = n_expected - n_resolved,
    epitope = "T"
  )
```

### Electrostatic surface complementarity (APBS)
```{bash, eval = FALSE}
### Electrostatic surface complementarity (APBS)
# Antibody-antigen binding depends on electrostatic complementarity at the
# epitope surface. Potential maps were generated externally as follows:
#
# 1. PDB2PQR (v3.7.1, AMBER force field) was run on each epitope-containing
#    structure to assign atomic charges and radii, generating a .pqr file
#    and APBS input file (.in).
# 2. APBS (v3.4.1) was run on each .in file to produce an electrostatic
#    potential grid (.dx file) for HER2 (1S78, 1N8Z), HER3 (1M6B),
#    HER4 (2AHX), and EGFR (1NQL).
# 3. Epitope Cα coordinates were exported from PyMOL for each structure.
#    IMPORTANT: coordinates must come from the UNALIGNED, original
#    structures — not from structures superimposed during the RMSD step —
#    since the APBS potential grid is defined in each structure's own
#    native coordinate frame.
# 4. The R functions below sample the potential grid at each epitope
#    Cα position (trilinear interpolation) and compare HER2 to each
#    family member via Pearson correlation.
```

```{r, eval=FALSE}
# DX maps generated externally via PDB2PQR + APBS; epitope coordinates exported
# from PyMOL in the original (unaligned) coordinate frame.

read_dx <- function(filepath) {
  lines <- readLines(filepath)
  grid_line <- grep("^object 1", lines, value = TRUE)
  dims <- as.integer(str_extract_all(grid_line, "\\d+")[[1]][2:4])
  
  num_pattern <- "-?[\\d]+\\.?[\\d]*[eE]?[+-]?\\d*"
  origin_line <- grep("^origin", lines, value = TRUE)
  origin <- as.numeric(str_match_all(origin_line, num_pattern)[[1]])
  
  delta_lines <- grep("^delta", lines, value = TRUE)
  dx <- as.numeric(str_match_all(delta_lines[1], num_pattern)[[1]])
  dy <- as.numeric(str_match_all(delta_lines[2], num_pattern)[[1]])
  dz <- as.numeric(str_match_all(delta_lines[3], num_pattern)[[1]])
  spacing <- c(dx[1], dy[2], dz[3])
  
  data_start <- max(grep("^object 3", lines)) + 1
  data_end <- grep("^attribute", lines)[1] - 1
  values <- as.numeric(unlist(strsplit(trimws(lines[data_start:data_end]), "\\s+")))
  
  list(dims = dims, origin = origin, spacing = spacing, values = values)
}

get_potential <- function(dx_map, x, y, z) {
  fi <- (x - dx_map$origin[1]) / dx_map$spacing[1]
  fj <- (y - dx_map$origin[2]) / dx_map$spacing[2]
  fk <- (z - dx_map$origin[3]) / dx_map$spacing[3]
  i <- floor(fi); j <- floor(fj); k <- floor(fk)
  fx <- fi - i; fy <- fj - j; fz <- fk - k
  
  if (i < 0 || i >= dx_map$dims[1]-1 ||
      j < 0 || j >= dx_map$dims[2]-1 ||
      k < 0 || k >= dx_map$dims[3]-1) return(NA)
  
  idx <- function(ii, jj, kk) ii * dx_map$dims[2] * dx_map$dims[3] + jj * dx_map$dims[3] + kk + 1
  
  v000 <- dx_map$values[idx(i, j, k)];     v100 <- dx_map$values[idx(i+1, j, k)]
  v010 <- dx_map$values[idx(i, j+1, k)];   v110 <- dx_map$values[idx(i+1, j+1, k)]
  v001 <- dx_map$values[idx(i, j, k+1)];   v101 <- dx_map$values[idx(i+1, j, k+1)]
  v011 <- dx_map$values[idx(i, j+1, k+1)]; v111 <- dx_map$values[idx(i+1, j+1, k+1)]
  
  v000*(1-fx)*(1-fy)*(1-fz) + v100*fx*(1-fy)*(1-fz) +
    v010*(1-fx)*fy*(1-fz) + v110*fx*fy*(1-fz) +
    v001*(1-fx)*(1-fy)*fz + v101*fx*(1-fy)*fz +
    v011*(1-fx)*fy*fz + v111*fx*fy*fz
}

# --- Pertuzumab APBS ---
pert_maps <- list(
  HER2 = read_dx(file.path(base_path, "1S78_her2.pqr.dx")),
  HER3 = read_dx(file.path(base_path, "1M6B_her3.pqr.dx")),
  HER4 = read_dx(file.path(base_path, "2AHX_her4.pqr.dx")),
  EGFR = read_dx(file.path(base_path, "1NQL_egfr.pqr.dx"))
)

pert_coords <- list(
  HER2 = read_csv(file.path(base_path, "her2_epitope_coords.csv"), show_col_types = FALSE),
  HER3 = read_csv(file.path(base_path, "her3_epitope_coords.csv"), show_col_types = FALSE),
  HER4 = read_csv(file.path(base_path, "her4_epitope_coords.csv"), show_col_types = FALSE),
  EGFR = read_csv(file.path(base_path, "egfr_epitope_coords.csv"), show_col_types = FALSE)
)

pert_potentials <- map2_dfr(names(pert_maps), names(pert_coords), function(prot, coord_name) {
  m <- pert_maps[[prot]]; c <- pert_coords[[coord_name]]
  c %>% rowwise() %>% mutate(protein = prot, potential = get_potential(m, x, y, z)) %>% ungroup()
})

her2_pot_pert <- pert_potentials %>% filter(protein == "HER2") %>% pull(potential)

pert_correlations <- map_dfr(c("HER3", "HER4", "EGFR"), function(prot) {
  other_pot <- pert_potentials %>% filter(protein == prot) %>% pull(potential)
  tibble(epitope = "P", protein = prot,
         correlation = cor(her2_pot_pert, other_pot, use = "complete.obs"),
         n_matched = length(other_pot))
})

# --- Trastuzumab APBS ---
tras_maps <- list(
  HER2 = read_dx(file.path(base_path, "1N8Z_her2.pqr.dx")),
  HER3 = read_dx(file.path(base_path, "1M6B_her3.pqr.dx")),
  HER4 = read_dx(file.path(base_path, "2AHX_her4.pqr.dx")),
  EGFR = read_dx(file.path(base_path, "1NQL_egfr.pqr.dx"))
)

tras_coords <- list(
  HER2 = read_csv(file.path(base_path, "tras_her2_epitope_coords.csv"), show_col_types = FALSE),
  HER3 = read_csv(file.path(base_path, "tras_her3_epitope_coords.csv"), show_col_types = FALSE),
  HER4 = read_csv(file.path(base_path, "tras_her4_epitope_coords.csv"), show_col_types = FALSE),
  EGFR = read_csv(file.path(base_path, "tras_egfr_epitope_coords.csv"), show_col_types = FALSE)
)

tras_potentials <- map2_dfr(names(tras_maps), names(tras_coords), function(prot, coord_name) {
  m <- tras_maps[[prot]]; c <- tras_coords[[coord_name]]
  c %>% rowwise() %>% mutate(protein = prot, potential = get_potential(m, x, y, z)) %>% ungroup()
}) %>%
  group_by(protein) %>%
  mutate(pos_idx = row_number()) %>%
  ungroup()

her2_pot_tras <- tras_potentials %>% filter(protein == "HER2") %>% select(pos_idx, potential)

tras_correlations <- map_dfr(c("HER3", "HER4", "EGFR"), function(prot) {
  other_pot <- tras_potentials %>% filter(protein == prot) %>% select(pos_idx, potential)
  matched <- inner_join(her2_pot_tras, other_pot, by = "pos_idx", suffix = c("_her2", "_other"))
  tibble(epitope = "T", protein = prot,
         correlation = cor(matched$potential_her2, matched$potential_other, use = "complete.obs"),
         n_matched = nrow(matched))
})

### Combined off-target summary table

apbs_combined <- bind_rows(pert_correlations, tras_correlations)
bfactor_combined <- bind_rows(pert_bfactor_summary, tras_bfactor_summary) %>%
  select(protein, epitope, mean_bfactor)

offtarget_summary <- full_summary %>%
  left_join(rmsd_table, by = c("protein", "epitope")) %>%
  left_join(bfactor_combined, by = c("protein", "epitope")) %>%
  left_join(apbs_combined %>% select(protein, epitope, correlation, n_matched),
            by = c("protein", "epitope")) %>%
  mutate(epitope = recode(epitope, "P" = "Pertuzumab", "T" = "Trastuzumab"))

print(offtarget_summary, width = Inf) 
```

# Session Info

```{r session_info}
sessionInfo()
```
