---
title: "AXL"

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_format = "html_document",
      output_dir = "../results/html_docs",
    )
  })
---

```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
setwd("../../")
source("cartcontent/scr/00_functions.R")
```

<h4>Description</h4>
Tyrosine-protein kinase receptor UFO ([AXL](https://www.uniprot.org/uniprot/P30530)). Receptor tyrosine kinase that transduces signals from the extracellular matrix into the cytoplasm by binding growth factor GAS6 and which is thus regulating many physiological processes including cell survival, cell proliferation, migration and differentiation.
<br> 

<h4>Clinical trials</h4> 
[NCT03393936](https://clinicaltrials.gov/ct2/show/NCT03393936) - Renal Cancer

<br>

<h4>Target gene expression</h4>
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 27, fig.height= 12}
setwd("../../")
boxplot_isoforms_all_tissues(hgnc = "AXL") # it creates the boxplot in the html and saves it as .pdf in the plot/isoform_expression folder
```
<br>

<h4>Target splice variants</h4>

<br>

<h4>Target essentiality</h4>

<br>

<h4>Protein properties</h4>
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 12, fig.height= 5}
setwd("../../")
plot_function("AXL") # it creates the plot in the html and saves it as .pdf in the plot/protein_properties folder
```