---
title: "CD70"

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_format = "html_document",
      output_dir = "../results/html_docs",
    )
  })
---

```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
setwd("../../")
source("cartcontent/scr/00_functions.R")
```

<br>

#### Description
CD70 antigen ([CD70](https://www.uniprot.org/uniprot/P32970)). Cytokine which is the ligand for CD27. The CD70-CD27 pathway plays an important role in the generation and maintenance of T cell immunity, in particular during antiviral responses.

<br> 

#### Clinical trials 
[NCT02830724](https://clinicaltrials.gov/ct2/show/NCT02830724) - Breast Cancer, Renal Cell Cancer, Ovarian Cancer, Pancreatic Cancer, Melanoma

<br>

#### Target gene expression
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 27, fig.height= 12}
setwd("../../")
boxplot_isoforms_all_tissues(hgnc = "CD70") # it creates the boxplot in the html and saves it as .pdf in the plot/isoform_expression folder
```
<br>

#### Target splice variants

<br>

#### Target essentiality

<br>

#### Protein properties
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 12, fig.height= 5}
setwd("../../")
plot_function("CD70") # it creates the plot in the html and saves it as .pdf in the plot/protein_properties folder
```