---
title: "EGFR"

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_format = "html_document",
      output_dir = "../results/html_docs",
    )
  })
---
<style>
  .scrollingplot{
      overflow-x:scroll;
      white-space: nowrap;
  }

  .scrollingplot img{
     max-width: 40cm;
  }

</style>

EVERYTHING TO BE UPDATED

```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
source("../scr/01_functions.R")

enst_expr <- read_rds("../results/01_wrangled_data/01_enst_expr.rds")
ensg_expr <- read_rds("../results/01_wrangled_data/01_ensg_expr.rds")
load(file = "../data/gte.Rdata") # transcripts for each gene + genomic coordinates
pheno_nh <- read_rds("../results/01_wrangled_data/01_pheno_no_haematological_cancers.rds")
hgnc_from_ensembl <- readRDS(file = "../results/00_hgnc_from_ensembl_biomart.rds")


list_of_categories <- unique(pheno_nh$TCGA_GTEX_main_category)

# find the gene in ensg_exp with the highest number of isoforms in order to
# decide the dimensions of the boxplot

l <- vector(mode = "numeric", length = nrow(ensg_expr))
for(i in 1:nrow(ensg_expr)) {
  ensg <- rownames(ensg_expr)[i]
  l[i] <- length(names(gte_list[[ensg]]))
  
  maxl <- (max(l))
  
}
```


<h4>Description</h4>
Epidermal growth factor receptor ([EGFR](https://www.uniprot.org/uniprot/P00533)). Receptor tyrosine kinase binding ligands of the EGF family and activating several signaling cascades to convert extracellular cues into appropriate cellular responses. 

<br> 

<h4>Clinical trials</h4> 
[NCT03542799](https://clinicaltrials.gov/ct2/show/NCT03542799) - Metastatic Colorectal Cancer <br>
[NCT01818323](https://clinicaltrials.gov/ct2/show/NCT01818323) - Head and Neck Cancer
[NCT03638206](https://clinicaltrials.gov/ct2/show/NCT03638206) - Hepatoma, Pancreatic Cancer, Colorectal Cancer, Esophagus Cancer, Lung Cancer, Melanoma, Ovarian Cancer, Renal Carcinoma <br> 
[NCT03941626](https://clinicaltrials.gov/ct2/show/NCT03941626)	- Esophagus Cancer, Hepatoma <br>
[NCT01454596](https://clinicaltrials.gov/ct2/show/NCT01454596)	- Glioblastoma Multiforme <br>
[NCT02664363](https://clinicaltrials.gov/ct2/show/NCT02664363)	- Glioblastoma Multiforme <br>
[NCT02844062](https://clinicaltrials.gov/ct2/show/NCT02844062)	- Glioblastoma Multiforme <br>
[NCT03267173](https://clinicaltrials.gov/ct2/show/NCT03267173)	- Pancreatic Cancer 

<br>

<h4>Target gene expression</h4>
The following plot describes the expression, in different tissues and TCGA cancers, of the different isoforms of the gene EGFR.
<div class="scrollingplot">
```{r , echo = FALSE, warning = FALSE, message = FALSE, fig.width = 80, fig.height = 40}
(plot <- boxplot_isoforms_all_tissues(hgnc = "EGFR", gte_list = gte_list, pheno = pheno_nh, enst_expr = enst_expr, hgnc_from_ensembl = hgnc_from_ensembl)) # it creates the boxplot in the html 
#save it as .pdf in the plot/isoform_expression folder
ggsave(filename = "EGFR_isoforms_expression_plot.pdf", plot = plot, device = "pdf",
         path = "../results/plots/isoform_expression/",
         width = 150, height = 100, units = "cm", limitsize = F)

```
</div>
<br>

<h4>Target essentiality</h4>
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 25, fig.height= 15}
setwd("../../")
target_essentiality_fun(gene_name_hgnc = "EGFR") # it creates the boxplot in the html and saves it as .pdf in the plot/target_essentiality folder
```

<br>

<h4>Protein properties</h4>
The following plot describes the target topology, highlighting the position of the endo-, ecto- and transmembrane domains, as well as the position of a signal peptide, when present. 
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 10, fig.height = 3}
setwd("../../")
plot_function("EGFR") # it creates the plot in the html and saves it as .pdf in the plot/protein_properties folder
```