---
title: "EGFRvIII"

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_format = "html_document",
      output_dir = "../results/html_docs",
    )
  })
---

```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
setwd("../../")
source("cartcontent/scr/00_functions.R")
```

<br>

#### Description
Epidermal growth factor receptor, variant 3 ([EGFRvIII](https://www.uniprot.org/uniprot/P00533)). Receptor tyrosine kinase binding ligands of the EGF family and activating several signaling cascades to convert extracellular cues into appropriate cellular responses.

<br> 

#### Clinical trials 
[NCT03638206](https://clinicaltrials.gov/ct2/show/NCT03638206) -	Hepatoma, Pancreatic Cancer, Colorectal Cancer, Esophagus Cancer, Lung Cancer, Melanoma, Ovarian Cancer, Renal Carcinoma <br> 
[NCT03941626](https://clinicaltrials.gov/ct2/show/NCT03941626)	- Esophagus Cancer, Hepatoma <br>
[NCT01454596](https://clinicaltrials.gov/ct2/show/NCT01454596)	- Glioblastoma Multiforme <br>
[NCT02664363](https://clinicaltrials.gov/ct2/show/NCT02664363)	- Glioblastoma Multiforme <br>
[NCT02844062](https://clinicaltrials.gov/ct2/show/NCT02844062)	- Glioblastoma Multiforme <br>
[NCT03267173](https://clinicaltrials.gov/ct2/show/NCT03267173)	- Pancreatic Cancer 

<br>

#### Target gene expression
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 27, fig.height= 12}
setwd("../../")
boxplot_isoforms_all_tissues(hgnc = "EGFRvIII") # it creates the boxplot in the html and saves it as .pdf in the plot/isoform_expression folder
```
<br>

#### Target splice variants

<br>

#### Target essentiality

<br>

#### Protein properties
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 12, fig.height= 5}
setwd("../../")
plot_function("EGFRvIII") # it creates the plot in the html and saves it as .pdf in the plot/protein_properties folder
```