---
title: "EPCAM"

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_format = "html_document",
      output_dir = "../results/html_docs",
    )
  })
---

```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
setwd("../../")
source("cartcontent/scr/00_functions.R")
```

<br>

#### Description
Epithelial cell adhesion molecule ([EPCAM](https://www.uniprot.org/uniprot/P16422)). May act as a physical homophilic interaction molecule between intestinal epithelial cells (IECs) and intraepithelial lymphocytes (IELs) at the mucosal epithelium for providing immunological barrier as a first line of defense against mucosal infection. Plays a role in embryonic stem cells proliferation and differentiation. Up-regulates the expression of FABP5, MYC and cyclins A and E.

<br> 

#### Clinical trials 
[NCT03013712](https://clinicaltrials.gov/ct2/show/NCT03013712) - Colon Cancer, Esophagus Cancer, Hepatic Carcinoma, Pancreatic Cancer, Prostate Cancer <br> 
[NCT02729493](https://clinicaltrials.gov/ct2/show/NCT02729493)	- Liver Neoplasms <br>
[NCT02725125](https://clinicaltrials.gov/ct2/show/NCT02725125)	- Stomach Neoplasms <br>

<br>

#### Target gene expression
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 27, fig.height= 12}
setwd("../../")
boxplot_isoforms_all_tissues(hgnc = "EPCAM") # it creates the boxplot in the html and saves it as .pdf in the plot/isoform_expression folder
```
<br>

#### Target splice variants

<br>

#### Target essentiality

<br>

#### Protein properties
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 12, fig.height= 5}
setwd("../../")
plot_function("EPCAM") # it creates the plot in the html and saves it as .pdf in the plot/protein_properties folder
```