---
title: "ERBB2"

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_format = "html_document",
      output_dir = "../results/html_docs",
    )
  })
---

```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
setwd("../../")
source("cartcontent/scr/00_functions.R")
```


<h4>Description</h4>
Receptor tyrosine-protein kinase erbB-2 ([ERBB2](https://www.uniprot.org/uniprot/P04626)). Protein tyrosine kinase that is part of several cell surface receptor complexes. In the nucleus it is involved in transcriptional regulation. Enhances protein synthesis and cell growth.

<br> 

<h4>Clinical trials</h4> 
[NCT03740256](https://clinicaltrials.gov/ct2/show/NCT03740256)	- Bladder Cancer, Breast Cancer, Colorectal    Cancer, Esophagus Cancer, Head and Neck Squamous Cell Carcinoma, Lung Cancer, Pancreatic Adenocarcinoma <br>
[NCT03696030](https://clinicaltrials.gov/ct2/show/NCT03696030)	- Breast Cancer <br>
[NCT02547961](https://clinicaltrials.gov/ct2/show/NCT02547961) -	Breast Cancer <br>
[NCT04511871](https://clinicaltrials.gov/ct2/show/NCT04511871) -	Breast Cancer, Ovarian Cancer, Sarcoma <br>
[NCT02713984](https://clinicaltrials.gov/ct2/show/NCT02713984)	- Breast Cancer, Colorectal Cancer, Lung Cancer, Ovarian Cancer, Pancreatic Adenocarcinoma <br>
[NCT01109095](https://clinicaltrials.gov/ct2/show/NCT01109095)	- Glioblastoma Multiforme <br>
[NCT03267173](https://clinicaltrials.gov/ct2/show/NCT03267173)	- Pancreatic Cancer <br>
[NCT00902044](https://clinicaltrials.gov/ct2/show/NCT00902044) -	Sarcoma 

<br>

<h4>Target gene expression</h4>
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 27, fig.height= 12}
setwd("../../")
boxplot_isoforms_all_tissues(hgnc = "ERBB2") # it creates the boxplot in the html and saves it as .pdf in the plot/isoform_expression folder
```
<br>

<h4>Target splice variants</h4>

<br>

<h4>Target essentiality</h4>

<br>

<h4>Protein properties</h4>
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 12, fig.height= 5}
setwd("../../")
plot_function("ERBB2") # it creates the plot in the html and saves it as .pdf in the plot/protein_properties folder
```