---
title: "FOLH1"

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_format = "html_document",
      output_dir = "../results/html_docs",
    )
  })
---
<style>
  .scrollingplot{
      overflow-x:scroll;
      white-space: nowrap;
  }

  .scrollingplot img{
     max-width: 40cm;
  }

</style>

```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
setwd("../../")
source("cartcontent/scr/01_functions.R")
```


<h4>Description</h4>
Glutamate carboxypeptidase 2 ([FOLH1](https://www.uniprot.org/uniprot/Q04609)) Has both folate hydrolase and N-acetylated-alpha-linked-acidic dipeptidase (NAALADase) activity. Has a preference for tri-alpha-glutamate peptides. In the intestine, required for the uptake of folate. In the brain, modulates excitatory neurotransmission through the hydrolysis of the neuropeptide, N-aceylaspartylglutamate (NAAG), thereby releasing glutamate. Involved in prostate tumor progression.

<br> 

<h4>Clinical trials</h4>
[NCT04633148](https://clinicaltrials.gov/ct2/show/NCT04633148) - Breast Cancer, Colorectal Cancer, Renal Cancer, Prostate Cancer <br>
[NCT03356795](https://clinicaltrials.gov/ct2/show/NCT03356795) - Cervical Cancer <br> 
[NCT04249947](https://clinicaltrials.gov/ct2/show/NCT04249947) - Prostate Cancer <br>
[NCT01140373](https://clinicaltrials.gov/ct2/show/NCT01140373) - Prostate Cancer <br>

<br>

<h4>Target gene expression</h4>
The following plot describes the expression, in different tissues and TCGA cancers, of the different isoforms of the gene FOLH1.
<div class="scrollingplot">
```{r , echo = FALSE, warning = FALSE, message = FALSE, fig.width = 80, fig.height = 40}
setwd("../../")
boxplot_isoforms_all_tissues(hgnc = "FOLH1") # it creates the boxplot in the html and saves it as .pdf in the plot/isoform_expression folder
```
</div>
<br>

<h4>Target essentiality</h4>
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 25, fig.height= 15}
setwd("../../")
target_essentiality_fun(gene_name_hgnc = "FOLH1") # it creates the boxplot in the html and saves it as .pdf in the plot/target_essentiality folder
```

<br>

<h4>Protein properties</h4>
The following plot describes the target topology, highlighting the position of the endo-, ecto- and transmembrane domains, as well as the position of a signal peptide, when present. 
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 10, fig.height = 3}
setwd("../../")
plot_function("FOLH1") # it creates the plot in the html and saves it as .pdf in the plot/protein_properties folder
```