---
title: "GPC3"

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_format = "html_document",
      output_dir = "../results/html_docs",
    )
  })
---

```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
setwd("../../")
source("cartcontent/scr/00_functions.R")
```

<br>

#### Description
Glypican-3 ([GPC3](https://www.uniprot.org/uniprot/P51654)). Cell surface proteoglycan that bears heparan sulfate. Used as a marker for hepatocellular carcinoma (HCC). When attached to the cell surface, stimulates the growth of HCC cells by increasing canonical Wnt signaling. 

<br> 

#### Clinical trials 
[NCT03884751](https://clinicaltrials.gov/ct2/show/NCT03884751) - Hepatocellular Carcinoma <br>
[NCT02905188](https://clinicaltrials.gov/ct2/show/NCT02905188) - Hepatocellular Carcinoma <br> 
[NCT04377932](https://clinicaltrials.gov/ct2/show/NCT04377932) - Hepatocellular Carcinoma <br>
[NCT02932956](https://clinicaltrials.gov/ct2/show/NCT02932956) - Hepatocellular Carcinoma <br>
[NCT04093648](https://clinicaltrials.gov/ct2/show/NCT04093648) - Hepatocellular Carcinoma <br>
[NCT02723942](https://clinicaltrials.gov/ct2/show/NCT02723942) - Hepatocellular Carcinoma <br>
[NCT04121273](https://clinicaltrials.gov/ct2/show/NCT04121273) - Hepatocellular Carcinoma <br>
[NCT03146234](https://clinicaltrials.gov/ct2/show/NCT03146234) - Hepatocellular Carcinoma <br>
[NCT03130712](https://clinicaltrials.gov/ct2/show/NCT03130712) - Hepatocellular Carcinoma <br>
[NCT02715362](https://clinicaltrials.gov/ct2/show/NCT02715362) - Hepatocellular Carcinoma <br>
[NCT03084380](https://clinicaltrials.gov/ct2/show/NCT03084380) - Hepatocellular Carcinoma <br>
[NCT02395250](https://clinicaltrials.gov/ct2/show/NCT02395250) - Hepatocellular Carcinoma <br>
[NCT02959151](https://clinicaltrials.gov/ct2/show/NCT02959151) - Hepatocellular Carcinoma <br>
[NCT02876978](https://clinicaltrials.gov/ct2/show/NCT02876978) - Lung Squamous Cell Carcinoma 

<br>

#### Target gene expression
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 27, fig.height= 12}
setwd("../../")
boxplot_isoforms_all_tissues(hgnc = "GPC3") # it creates the boxplot in the html and saves it as .pdf in the plot/isoform_expression folder
```
<br>

#### Target splice variants

<br>

#### Target essentiality

<br>

#### Protein properties
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 12, fig.height= 5}
setwd("../../")
plot_function("GPC3") # it creates the plot in the html and saves it as .pdf in the plot/protein_properties folder
```