---
title: "MET"

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_format = "html_document",
      output_dir = "../results/html_docs",
    )
  })
---

```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
setwd("../../")
source("cartcontent/scr/00_functions.R")
```

<br>

#### Description
Hepatocyte growth factor receptor ([MET](https://www.uniprot.org/uniprot/P08581)). Receptor tyrosine kinase that transduces signals from the extracellular matrix into the cytoplasm by binding to hepatocyte growth factor/HGF ligand. 

<br> 

#### Clinical trials 
[NCT03060356](https://clinicaltrials.gov/ct2/show/NCT03060356) - Breast Cancer, Melanoma <br>
[NCT03638206](https://clinicaltrials.gov/ct2/show/NCT03638206) - Colorectal Cancer, Esophagus Cancer, Renal Carcinoma, Hepatoma, Lung Cancer, Mesothelioma, Ovarian Cancer, Pancreatic Cancer, Melanoma <br>
[NCT03672305](https://clinicaltrials.gov/ct2/show/NCT03672305) - Hepatoma

<br>

#### Target gene expression
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 27, fig.height= 12}
setwd("../../")
boxplot_isoforms_all_tissues(hgnc = "MET") # it creates the boxplot in the html and saves it as .pdf in the plot/isoform_expression folder
```
<br>

#### Target splice variants

<br>

#### Target essentiality

<br>

#### Protein properties
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 12, fig.height= 5}
setwd("../../")
plot_function("MET") # it creates the plot in the html and saves it as .pdf in the plot/protein_properties folder
```