---
title: "MUC1"

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_format = "html_document",
      output_dir = "../results/html_docs",
    )
  })
---

```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
setwd("../../")
source("cartcontent/scr/00_functions.R")
```

<br>

#### Description
Mucin-1 ([MUC1](https://www.uniprot.org/uniprot/P15941)). Can act both as an adhesion and an anti-adhesion protein. May provide a protective layer on epithelial cells against bacterial and enzyme attack. Promotes tumor progression.

<br> 

#### Clinical trials 
[NCT03356795](https://clinicaltrials.gov/ct2/show/NCT03356795) - Cervical Cancer <br>
[NCT02839954](https://clinicaltrials.gov/ct2/show/NCT02839954)	- Colorectal Carcinoma, Hepatocellular Carcinoma, Pancreatic Carcinoma <br>
[NCT02617134](https://clinicaltrials.gov/ct2/show/NCT02617134)	- Colorectal Carcinoma <br>
[NCT02587689](https://clinicaltrials.gov/ct2/show/NCT02587689)	- Hepatocellular Carcinoma, Pancreatic Carcinoma <br>
[NCT04025216](https://clinicaltrials.gov/ct2/show/NCT04025216) -	Ovarian Cancer <br>
[NCT03267173](https://clinicaltrials.gov/ct2/show/NCT03267173)	- Pancreatic Carcinoma <br>

<br>

#### Target gene expression
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width =50, fig.height= 12}
setwd("../../")
boxplot_isoforms_all_tissues(hgnc = "MUC1") # it creates the boxplot in the html and saves it as .pdf in the plot/isoform_expression folder
```
<br>

#### Target splice variants

<br>

#### Target essentiality

<br>

#### Protein properties
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 12, fig.height= 5}
setwd("../../")
plot_function("MUC1") # it creates the plot in the html and saves it as .pdf in the plot/protein_properties folder
```