---
title: "TNFRSF17"

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_format = "html_document",
      output_dir = "../results/html_docs",
    )
  })
---

```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
setwd("../../")
source("cartcontent/scr/00_functions.R")
```

<br>

#### Description
Tumor necrosis factor receptor superfamily member 17 ([TNFRSF17](https://www.uniprot.org/uniprot/Q02223)). Receptor for TNFSF13B/BLyS/BAFF and TNFSF13/APRIL. Promotes B-cell survival and plays a role in the regulation of humoral immunity. Activates NF-kappa-B and JNK.

<br> 

#### Clinical trials 
[NCT03302403](https://clinicaltrials.gov/ct2/show/NCT03302403) - Hepatocellular Carcinoma, Pancreatic Cancer

<br>

#### Target gene expression
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 27, fig.height= 12}
setwd("../../")
boxplot_isoforms_all_tissues(hgnc = "TNFRSF17") # it creates the boxplot in the html and saves it as .pdf in the plot/isoform_expression folder
```
<br>

#### Target splice variants

<br>

#### Target essentiality

<br>

#### Protein properties
```{r, eval = TRUE, echo = FALSE, warning = FALSE, message = F, fig.width = 12, fig.height= 5}
setwd("../../")
plot_function("TNFRSF17") # it creates the plot in the html and saves it as .pdf in the plot/protein_properties folder
```